# Bayesian t-test


The following is based on Kruschke's 2012 JEP article 'Bayesian estimation supersedes the t-test (BEST)' with only minor changes to Stan model. It uses the JAGS/BUGS code in the paper's Appendix B as the reference.              

## Data Setup

Create two groups of data. Play around with the specs if you like.

```{r bayes-t-setup}
library(tidyverse)

set.seed(1234)

N_g   = 2       # N groups
N_1   = 50      # N for group 1
N_2   = 50      # N for group 2
mu_1  = 1       # mean for group 1
mu_2  = -.5     # mean for group 1
sigma_1 = 1     # sd for group 1
sigma_2 = 1     # sd for group 1

y_1 = rnorm(N_1, mu_1, sigma_1)
y_2 = rnorm(N_2, mu_2, sigma_2)
y   = c(y_1, y_2)

group_id = as.numeric(gl(2, N_1))

# if unbalanced
# group = 1:2
# group_id = rep(group, c(N_1,N_2))

d = data.frame(y, group_id)

tidyext::num_by(d, y, group_id)  # personal package, not necessary
```



## Model Code

The Stan code.  

```{stan bayes-t-test, output.var='bayes_t_test'}
data {
  int<lower = 1> N;                              // sample size 
  int<lower = 2> N_g;                            // number of groups
  vector[N] y;                                   // response
  int<lower = 1, upper = N_g> group_id[N];       // group ID
}

transformed data{
  real y_mean;                                   // mean of y; see mu prior
  
  y_mean = mean(y); 
}

parameters {
  vector[2] mu;                                  // estimated group means and sd
  vector<lower = 0>[2] sigma;                    // Kruschke puts upper bound as well; ignored here
  real<lower = 0, upper = 100> nu;               // df for t distribution
}

model {
  // priors
  // note that there is a faster implementation of this for stan, 
  // and that the sd here is more informative than in Kruschke paper
  mu    ~ normal(y_mean, 10);                       
  sigma ~ cauchy(0, 5);
  
  // Based on Kruschke; makes average nu 29 
  // might consider upper bound, as if too large then might as well switch to normal
  nu    ~ exponential(1.0/29);                
  
  // likelihood
  for (n in 1:N) {
    y[n] ~ student_t(nu, mu[group_id[n]], sigma[group_id[n]]);
    
    // compare to normal; remove all nu specifications if you do this;
    //y[n] ~ normal(mu[group_id[n]], sigma[group_id[n]]);           
  }
}

generated quantities {
  vector[N] y_rep;                               // posterior predictive distribution
  real mu_diff;                                  // mean difference
  real cohens_d;                                 // effect size; see footnote 1 in Kruschke paper
  real CLES;                                     // common language effect size
  real CLES2;                                    // a more explicit approach; the mean should roughly equal CLES

  for (n in 1:N) {
    y_rep[n] = student_t_rng(nu, mu[group_id[n]], sigma[group_id[n]]);
  }

  mu_diff  = mu[1] - mu[2];
  cohens_d = mu_diff / sqrt(sum(sigma)/2);
  CLES     = normal_cdf(mu_diff / sqrt(sum(sigma)), 0, 1);
  CLES2    = student_t_rng(nu, mu[1], sigma[1]) - student_t_rng(nu, mu[2], sigma[2]) > 0;
}
```


## Estimation

Run the model.

```{r bayes-t-est, results='hide'}
stan_data = list(
  N   = length(y),
  N_g = N_g,
  group_id = group_id,
  y = y
)

library(rstan) 

fit = sampling(
  bayes_t_test,
  data = stan_data,
  thin = 4
)


```


## Comparison

Let's take a look.

```{r bayes-t-results}
print(
  fit,
  pars   = c('mu', 'sigma', 'mu_diff', 'cohens_d', 'CLES', 'CLES2', 'nu'),
  probs  = c(.025, .5, .975), 
  digits = 3
)
```


Now we extract quantities of interest for more processing/visualization. Compare population and observed data values to estimates in summary to the observed mean difference.

```{r bayes-t-extract}
y_rep   = extract(fit, par = 'y_rep')$y_rep
mu_diff = extract(fit, par = 'mu_diff')$mu_diff

init = d %>% 
  group_by(group_id) %>% 
  summarise(
    mean = mean(y),
    sd = sd(y),
  )

means = init$mean
sds   = init$sd

mu_1 - mu_2           # based on population values
abs(diff(means))      # observed in data
```

Compare estimated [Cohen's d](https://en.wikipedia.org/wiki/Effect_size#Cohen's_d).

```{r bayes-t-cohens-d}
cohens_d = extract(fit, par='cohens_d')$cohens_d
(mu_1 - mu_2) / sqrt((sigma_1 ^ 2 + sigma_2 ^ 2)/2)      # population
(means[1] - means[2]) / sqrt(sum(sds^2)/2)               # observed
```


[Common language effect size](https://en.wikipedia.org/wiki/Effect_size#Common_language_effect_size) is the probability that a randomly selected score from one population will be greater than a randomly sampled score from the other.

```{r bayes-t-cohens-CLES}
CLES = extract(fit, par='CLES')$CLES
pnorm((mu_1 - mu_2) / sqrt(sigma_1^2 + sigma_2^2))       # population
pnorm((means[1] - means[2]) / sqrt(sum(sds^2)))          # observed
```

Compare to Welch's t-test that does not assume equal variances.

```{r bayes-t-welch}
t.test(y_1, y_2)
```


Compare to <span class="pack" style = "">BEST</span>. Note that it requires  <span class="pack" style = "">coda</span>, whose <span class="func" style = "">traceplot</span> function will mask <span class="pack" style = "">rstan's</span>.

```{r bayes-t-BEST}
library(BEST)

BESTout = BESTmcmc(
  y_1,
  y_2,
  numSavedSteps = 10000,
  thinSteps = 10,
  burnInSteps = 2000
)

summary(BESTout)
```


## Visualization

We can plot posterior predictive distribution vs. observed data density.

```{r bayes-t-pp-check}
gdat = y_rep %>% 
  as.data.frame() %>% 
  pivot_longer(everything(), names_to =  'iteration') %>% 
  mutate(iteration   = as.integer(str_extract(iteration, '[0-9]+')),
         observation = factor(rep(1:nrow(y_rep), e = 100)))


library(bayesplot)

pp_check(
  stan_data$y,
  rstan::extract(fit, par = 'y_rep')$y_rep[1:10, ], 
  fun = 'dens_overlay'
)
```

We can expand this to incorporate the separate groups and observed values. Solid lines and dots represent the observed data.

```{r bayes-t-vis, echo = FALSE}
# change e to match your number of posterior draws
gdat = gdat %>% 
  mutate(group_id = factor(rep(group_id, e = 400)))

ggplot(aes(x = value), data = gdat) +
  geom_density(aes(group = group_id, fill = group_id),
               color = NA,
               alpha = .25) +
  geom_line(aes(group = observation, color = group_id),
            stat  = 'density',
            alpha = .05) +
  geom_point(
    aes(
      x = y,
      y = 0,
      color = factor(group_id)
    ),
    alpha = .15,
    size = 5,
    data = data.frame(y, group_id)
  ) +
  geom_density(aes(group = group_id, color = group_id, x = y),
               alpha = .05,
               data.frame(group_id = factor(group_id), y)) +
  scico::scale_color_scico_d(end = .6, aesthetics = c('color', 'fill')) +
  xlim(c(-8, 8))   # might get a warning if extreme values are cut out


### plot mean difference or other values of interest
ggplot(aes(x = mu_diff), data = data.frame(mu_diff = mu_diff)) +
  geom_density(alpha = .25, color = 'gray92') +
  geom_point(x = mu_diff,
             y = 0,
             alpha = .01,
             size = 3) +
  geom_path(
    aes(x = quantile(mu_diff, c(.025, .975)), y = c(.2, .2)),
    size = 2,
    alpha = .5,
    color = '#b2001d',
    data = data.frame()
  ) +
  xlim(c(0, 3.5)) +
  labs(x = 'μ1 - μ2')
```

Plots from the BEST model.

```{r  bayes-t-vis-Best, echo=2}
par(mfrow = c(2, 2))
walk(c("mean", "sd", "effect", "nu"), function(p) plot(BESTout, which = p))
layout(1)
```



## Source

Original code available at:
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/Bayesian/rstant_testBEST.R