# Bayesian Multilevel Mediation

The following demonstrates an indirect effect in a multilevel situation.  It is
based on Yuan & MacKinnon 2009, which provides some Bugs code.  In what follows
we essentially have two models, one where the 'mediator' is the response; the
other regards the primary response of interest (noted `y`).  They will be referred
to with Med or Main respectively.  


## Data Setup 

The two main models are expressed conceptually as follows:

$$\textrm{Mediator} \sim \alpha_{Med} + \beta_{Med}\cdot X$$

$$y \sim \alpha_{Main} + \beta_{1\_{Main}}\cdot X + \beta_{2\_{Main}}\cdot \textrm{Mediator}$$

However, there will be random effects for a grouping variable for each coefficient, i.e. random intercepts and slopes, for both the mediator model and the outcome model.

```{r bayes-med-prelim, echo=FALSE}
N  = 1000
n_groups = 50
n_per_group = N/n_groups
```


Let's create data to this effect. In the following we will ultimately have `r N` total observations, with `r n_groups` groups (`r n_per_group` observations each).

```{r bayes-med-setup}
library(tidyverse)

set.seed(8675309)

N  = 1000
n_groups = 50
n_per_group = N/n_groups

# random effects for mediator model
# create cov matrix of RE etc. with no covariance between model random effects
# covmat_RE = matrix(c(1,-.15,0,0,0,
#                       -.15,.4,0,0,0,
#                       0,0,1,-.1,.15,
#                       0,0,-.1,.3,0,
#                       0,0,.15,0,.2), nrow=5, byrow = T)

# or with slight cov added to indirect coefficient RE; both matrices are pos def
covmat_RE = matrix(c( 1.00, -0.15,  0.00,  0.00,  0.00,
                     -0.15,  0.64,  0.00,  0.00, -0.10,
                      0.00,  0.00,  1.00, -0.10,  0.15,
                      0.00,  0.00, -0.10,  0.49,  0.00,
                      0.00, -0.10,  0.15,  0.00,  0.25), nrow = 5, byrow = TRUE)

# inspect
covmat_RE

# inspect as correlation
cov2cor(covmat_RE)
colnames(covmat_RE) = rownames(covmat_RE) = 
  c('alpha_Med', 'beta_Med', 'alpha_Main', 'beta1_Main', 'beta2_Main')

# simulate random effects
re = MASS::mvrnorm(
  n_groups,
  mu    = rep(0, 5),
  Sigma = covmat_RE,
  empirical = TRUE
)

# random effects for mediator model
ranef_alpha_Med = rep(re[, 'alpha_Med'], e = n_per_group)
ranef_beta_Med  = rep(re[, 'beta_Med'],  e = n_per_group)

# random effects for main model                                                 
ranef_alpha_Main = rep(re[, 'alpha_Main'], e = n_per_group)
ranef_beta1_Main = rep(re[, 'beta1_Main'], e = n_per_group)
ranef_beta2_Main = rep(re[, 'beta2_Main'], e = n_per_group)

## fixed effects
alpha_Med = 2
beta_Med  = .2

alpha_Main = 1
beta1_Main = .3
beta2_Main = -.2

# residual variance
resid_Med  = MASS::mvrnorm(N, 0, .75^2, empirical = TRUE)
resid_Main = MASS::mvrnorm(N, 0, .50^2, empirical = TRUE)


# Collect parameters for later comparison
params = c(
  alpha_Med  = alpha_Med,
  beta_Med   = beta_Med,
  sigma_Med  = sd(resid_Med),
  alpha_Main = alpha_Main,
  beta1_Main = beta1_Main,
  beta2_Main = beta2_Main,
  sigma_y    = sd(resid_Main),
  alpha_Med_sd = sqrt(diag(covmat_RE)[1]),
  beta_Med_sd  = sqrt(diag(covmat_RE)[2]),
  alpha_sd = sqrt(diag(covmat_RE)[3]),
  beta1_sd = sqrt(diag(covmat_RE)[4]),
  beta2_sd = sqrt(diag(covmat_RE)[5])
)

ranefs =  cbind(
  gamma_alpha_Med = unique(ranef_alpha_Med),
  gamma_beta_Med  = unique(ranef_beta_Med),
  gamma_alpha = unique(ranef_alpha_Main),
  gamma_beta1 = unique(ranef_beta1_Main),
  gamma_beta2 = unique(ranef_beta2_Main)
)
```

Finally, we can create the data for analysis.  

```{r bayes-med-setup2}
X = rnorm(N, sd = 2)

Med = (alpha_Med + ranef_alpha_Med) + (beta_Med + ranef_beta_Med) * X + resid_Med[, 1]

y = (alpha_Main + ranef_alpha_Main) + (beta1_Main + ranef_beta1_Main) * X + 
  (beta2_Main + ranef_beta2_Main) *  Med + resid_Main[, 1]

group = rep(1:n_groups, e = n_per_group)

standat = list(
  X   = X,
  Med = Med,
  y   = y,
  Group = group,
  J = length(unique(group)),
  N = length(y)
)
```


## Model Code


In the following, the cholesky decomposition of the RE covariance matrix is used for efficiency.  As a rough guide, the default data with `r `N` observations took about a minute or so to run.



```{stan bayes-med-model, output.var='bayes_med_model'}
data {
  int<lower = 1> N;                              // Sample size
  vector[N] X;                                   // Explanatory variable
  vector[N] Med;                                 // Mediator
  vector[N] y;                                   // Response
  int<lower = 1> J;                              // Number of groups
  int<lower = 1,upper = J> Group[N];             // Groups
}

parameters{
  real alpha_Med;                                // mediator model reg parameters and related
  real beta_Med;
  real<lower = 0> sigma_alpha_Med;
  real<lower = 0> sigma_beta_Med;
  real<lower = 0> sigmaMed;

  real alpha_Main;                               // main model reg parameters and related
  real beta1_Main;
  real beta2_Main;
  real<lower = 0> sigma_alpha;
  real<lower = 0> sigma_beta1;
  real<lower = 0> sigma_beta2;
  real<lower = 0> sigma_y;

  cholesky_factor_corr[5] Omega_chol;            // chol decomp of corr matrix for random effects

  vector<lower = 0>[5] sigma_ranef;              // sd for random effects

  matrix[J,5] gamma;                             // random effects
}

transformed parameters{
  vector[J] gamma_alpha_Med;
  vector[J] gamma_beta_Med;

  vector[J] gamma_alpha;
  vector[J] gamma_beta1;
  vector[J] gamma_beta2;

  for (j in 1:J){
    gamma_alpha_Med[j] = gamma[j,1];
    gamma_beta_Med[j]  = gamma[j,2];
    gamma_alpha[j] = gamma[j,3];
    gamma_beta1[j] = gamma[j,4];
    gamma_beta2[j] = gamma[j,5];
  }
}

model {
  vector[N] mu_y;                                // linear predictors for response and mediator
  vector[N] mu_Med;
  matrix[5,5] D;
  matrix[5,5] DC;

  // priors
  // mediator model
  // fixef
  // for scale params the cauchy is a little more informative here due 
  // to the nature of the data
  sigma_alpha_Med ~ cauchy(0, 1);                
  sigma_beta_Med  ~ cauchy(0, 1);
  alpha_Med ~ normal(0, sigma_alpha_Med);   
  beta_Med  ~ normal(0, sigma_beta_Med);

  // residual scale
  sigmaMed ~ cauchy(0, 1);

  // main model
  // fixef
  sigma_alpha ~ cauchy(0, 1);
  sigma_beta1 ~ cauchy(0, 1);
  sigma_beta2 ~ cauchy(0, 1);
  alpha_Main  ~ normal(0, sigma_alpha);      
  beta1_Main  ~ normal(0, sigma_beta1);
  beta2_Main  ~ normal(0, sigma_beta2);

  // residual scale
  sigma_y ~ cauchy(0, 1);

  // ranef sampling via cholesky decomposition
  sigma_ranef ~ cauchy(0, 1);
  Omega_chol  ~  lkj_corr_cholesky(2.0);

  D  = diag_matrix(sigma_ranef);
  DC = D * Omega_chol;
  
  for (j in 1:J)                                 // loop for Group random effects
    gamma[j] ~ multi_normal_cholesky(rep_vector(0, 5), DC);

  // Linear predictors
  for (n in 1:N){
    mu_Med[n] = alpha_Med + gamma_alpha_Med[Group[n]] + 
    (beta_Med + gamma_beta_Med[Group[n]]) * X[n];
    
    mu_y[n]   = alpha_Main + gamma_alpha[Group[n]] + 
    (beta1_Main + gamma_beta1[Group[n]]) * X[n] + 
    (beta2_Main + gamma_beta2[Group[n]]) * Med[n] ;
  }
  
  
  // sampling for primary models
  Med ~ normal(mu_Med, sigmaMed);
  y   ~ normal(mu_y, sigma_y);
}

generated quantities{
  real naive_ind_effect;
  real avg_ind_effect;
  real total_effect;
  matrix[5,5] cov_RE;
  vector[N] y_hat;    
  
  cov_RE = diag_matrix(sigma_ranef) * tcrossprod(Omega_chol) * diag_matrix(sigma_ranef);

  naive_ind_effect = beta_Med*beta2_Main;
  avg_ind_effect   = beta_Med*beta2_Main + cov_RE[2,5];    // add cov of random slopes for mediator effects
  total_effect     = avg_ind_effect + beta1_Main; 
  
  for (n in 1:N){
    y_hat[n]   = alpha_Main + gamma_alpha[Group[n]] + 
    (beta1_Main + gamma_beta1[Group[n]]) * X[n] + 
    (beta2_Main + gamma_beta2[Group[n]]) * Med[n] ;
  }
}
```



## Estimation

Run the model and examine results.  The following assumes a character string or file (`bayes_med_model`) of the previous model code.

```{r bayes-med-est, cache.rebuild=F}
library(rstan)

fit = sampling(
  bayes_med_model,
  data    = standat,
  iter    = 3000,
  warmup  = 2000,
  thin    = 4,
  cores   = 4,
  control = list(adapt_delta = .99, max_treedepth = 15)
)
```




## Comparison

Main parameters include fixed and random effect standard deviation, plus those related to indirect effect.

```{r bayes-med-comp1}
mainpars = c(
  'alpha_Med',
  'beta_Med',
  'sigmaMed',
  'alpha_Main',
  'beta1_Main',
  'beta2_Main',
  'sigma_y',
  'sigma_ranef',
  'naive_ind_effect',
  'avg_ind_effect',
  'total_effect'
)

print(
  fit,
  digits = 3,
  probs  = c(.025, .5, 0.975),
  pars   = mainpars
)
```



We can use a piecemeal mixed model via <span class='pack'>lme4</span> for initial comparison.  However, it can't directly estimate mediated effect, and it won't pick up on correlation of random effects between models.

```{r bayes-med-comp2}
library(lme4)
mod_Med = lmer(Med ~ X + (1 + X | group))
summary(mod_Med)

mod_Main = lmer(y ~ X + Med + (1 + X + Med | group))
summary(mod_Main)

# should equal the naive estimate in the following code
lme_indirect_effect = fixef(mod_Med)['X'] * fixef(mod_Main)['Med']
```



Using the <span class='pack'>mediation</span> package will provide a better estimate, and can handle this simple mixed model setting.

```{r bayes-med-comp3}
# library(mediation)

mediation_mixed = mediation::mediate(
  model.m  = mod_Med,
  model.y  = mod_Main,
  treat    = 'X',
  mediator = 'Med'
)

summary(mediation_mixed)
```



Extract parameters for comparison.

```{r bayes-med-comapre-extract}
pars_primary = get_posterior_mean(fit, pars = mainpars)[, 5]
pars_re_cov  = get_posterior_mean(fit, pars = 'Omega_chol')[, 5] # or take 'cov_RE' from monte carlo sim
pars_re      = get_posterior_mean(fit, pars = c('sigma_ranef'))[, 5]
```


Fixed effects and random effect variances.

```{r bayes-med-compare-fe-re-var, echo = FALSE}
tibble(
  param = names(params),
  true  = params,
  bayes = pars_primary[1:12],
  lme4  = c(
    mod_Med@beta,
    summary(mod_Med)$sigma,
    mod_Main@beta,
    summary(mod_Main)$sigma,
    summary(mod_Med)$varcor$group[1, 1]^.5,
    summary(mod_Med)$varcor$group[2, 2]^.5,
    summary(mod_Main)$varcor$group[1, 1]^.5,
    summary(mod_Main)$varcor$group[2, 2]^.5,
    summary(mod_Main)$varcor$group[3, 3]^.5
  )
) %>% 
  kable_df()
```



Compare the covariances of the random effects. The first shows the full covariance matrix for mediator and outcome, then broken out separately.

```{r bayes-med-compare-re-cov, echo = FALSE}
covmat_RE_est = diag(pars_re) %*% 
  tcrossprod(matrix(pars_re_cov, ncol = 5, byrow = T)) %*%
  diag(pars_re)

list(true = covmat_RE, estimates = round(covmat_RE_est, 2))

vcov_Med = covmat_RE_est[1:2, 1:2]

list(
  vcov_Med = covmat_RE[1:2, 1:2],
  vcov_Med_bayes = round(vcov_Med, 2),
  vcov_Med_lme4  = round(summary(mod_Med)$varcor$group[1:2, 1:2], 2)
)

vcov_Main = covmat_RE_est[3:5, 3:5]
list(
  vcov_Main = covmat_RE[3:5, 3:5],
  vcov_Main_bayes = round(vcov_Main, 2),
  vcov_Main_lme4  = round(summary(mod_Main)$varcor$group[1:3, 1:3], 2)
)
```



Compare indirect effects

```{r bayes-med-compare-ind, echo=FALSE}
tibble(
  true = beta_Med * beta2_Main + covmat_RE[2, 5],
  est_bayes   = get_posterior_mean(fit, 'avg_ind_effect')[, 5],
  naive_bayes = get_posterior_mean(fit, 'naive_ind_effect')[, 5],
  naive_lmer  = lme_indirect_effect,
  mediation_pack = mediation_mixed$d0
) %>% 
  kable_df()
```

Note that you can use <span class="pack" style = "">brms</span> to estimate this model as follows.  The `i` allows the random effects to correlate across Mediator and outcome models.  We have to convert the correlation estimate back to the covariance estimate to get the indirect value to compare to our base Stan result.

```{r bayes-med-brms, results='hide'}
library(brms)

f = 
  bf(Med ~ X + (1 + X |i| group)) +
  bf(y   ~ X + Med + (1 + X + Med |i| group)) +
  set_rescor(FALSE)
       

fit_brm = brm(
  f,
  data  = data.frame(X, Med, y, group),
  cores = 4,
  thin  = 4,
  seed  = 1234,
  control = list(adapt_delta = .99, max_treedepth = 15)
)
```


```{r bayes-med-brms-show}
summary(fit_brm)

hypothesis(
  fit_brm,
  'b_y_Med*b_Med_X + cor_group__Med_X__y_Med*sd_group__Med_X*sd_group__y_Med = 0',
  class = NULL,
  seed  =  1234
)
```


## Visualization

```{r bayes-med-vis}
library(bayesplot)

pp_check(
  standat$y, 
  rstan::extract(fit, par = 'y_hat')$y_hat[1:10, ], 
  fun = 'dens_overlay'
)
```

## Source

Original code available at:
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/Bayesian/rstan_multilevelMediation.R




