# Topic Model

An implementation of Gibbs sampling for topic models for the example in  section
4 of [Steyvers and Griffiths (2007)](http://cocosci.berkeley.edu/tom/papers/SteyversGriffiths.pdf). A very clear intro in my opinion. The core of the functions's code comprises mostly cosmetic
changes to that found [here](https://bcomposes.wordpress.com/2012/05/29/gibbs-sampler-for-toy-topic-model-example/). Added are the creation of a function with several arguments, plotting etc.



## Data Setup

```{r tm-setup}
library(tidyverse)

vocab = factor(c("river", "stream", "bank", "money", "loan"))
K = 2  # n of topics
v = length(vocab)  # number of unique words
d = 16 # number of documents 
```



Topic 1 gives equal probability to money loan and bank (zero for river and stream).
Topic 2 gives equal probability to river stream and bank (zero for money and loan).

Next we create a document term matrix. Each doc consists of a mix of 16 tokens of the vocab-
the first few regard financial banks, the last several water banks, and the docs in between possess a mixed vocab.

```{r tm-setup2}
dtm = matrix(c(0,0,4,6,6,
               0,0,5,7,4,
               0,0,7,5,4,
               0,0,7,6,3,
               0,0,7,2,7,
               0,0,9,3,4,
               1,0,4,6,5,
               1,2,6,4,3,
               1,3,6,4,2,
               2,3,6,1,4,
               2,3,7,3,1,
               3,6,6,1,0,
               6,3,6,0,1,
               2,8,6,0,0,
               4,7,5,0,0,
               5,7,4,0,0), ncol = v, byrow = TRUE)

row.names(dtm) = paste0('doc', 1:d)
colnames(dtm) = vocab
```


Next we create additional aspects to initialize the setup.

```{r tm-setup3}
# matrix of words in each document
wordmat = t(apply(dtm, 1, function(row) rep(vocab, row)))

# initialize random topic assignments to words
T0 = apply(wordmat, c(1, 2),  function(token) sample(1:2, 1))

# word by topic matrix of counts containing the number of times word w is
# assigned to topic j
C_wt = sapply(vocab, function(word)  cbind(sum(T0[wordmat == word] == 1),
                                           sum(T0[wordmat == word] == 2)))

C_wt = t(C_wt)
row.names(C_wt) = vocab


# topic by document matrix of counts containing the number of times topic j is
# assigned to a word in document d
C_dt = t(apply(T0, 1, table))
```




## Function

Note that this function is not self contained in that it uses some of the objects created above, assuming they are in the global environment. It has the capacity for a pseudo-visualization which can be instructive, but isn't shown for this document (and it requries <span class="pack" style = "">corrplot</span>), and I haven't tested it recently.

```{r tm-function}
topic_model = function(
  alpha,
  beta,
  nsim    = 2000,
  warmup  = nsim / 2,
  thin    = 10,
  verbose = TRUE,
  plot    = FALSE,
  dotsortext = 'dots'
) {
  
  ### Arguments: 
  # alpha and beta: hyperparameters
  # nsim: the total number of simulations
  # warmup: the number of initial sims to discard
  # thin: keep every thin simulation after warmup 
  # verbose: to print every 100th iteration and total time
  # plot: to visualize every x sim; the plot will show current estimates of
  # theta, phi, and topic assignments for each term in the docs; can be
  # visualized as dots or the actual terms; for the latter you may need to
  # fiddle with your viewing area size so that terms don't appear to run
  # together; plotting will increase runtime
  
  ### requires corrplot and abind packages.
  
  # initialize 
  Z = T0                                                   # topic assignments
  saveSim   = seq(warmup + 1, nsim, thin)                  # iterations to save
  thetaList = list()                                       # saved theta estimates
  phiList   = list()                                       # saved phi estimates
  p = proc.time()
  
  # for every simulation
  for (s in 1:nsim) {
    if(verbose && s %% 100 == 0) {                         # paste every 100th iteration if desired
      secs = round((proc.time() - p)[3],2)
      min  = round(secs/60, 2)
      message(paste0(
        'Iteration number: ',
        s,
        '\n',
        'Total time: ',
        ifelse(secs <= 60, paste(secs, 'seconds'), paste(min, 'minutes'))
      ))
    }
    
    # Visualization
    if(plot > 0 && s >1 &&  s %% plot == 0) {                # plot every value of argument
      require(corrplot)
      
      layout(matrix(c(1, 1, 2, 2, 3, 3, 3, 3, 3, 3), ncol = 10))
      corrplot(
        theta,
        is.corr = FALSE,
        method  = 'color',
        tl.cex  = .75,
        tl.col  = 'gray50',
        cl.pos  = 'n',
        addgrid = NA
      )
      corrplot(
        phi,
        is.corr = FALSE,
        method  = 'color',
        tl.cex  = 1,
        tl.col  = 'gray50',
        cl.pos  = 'n',
        addgrid = NA
      )
      if (dotsortext == 'dots') {
        Zplot = Z
        Zplot[Zplot == 2] = -1
        corrplot(
          Zplot,
          is.corr = FALSE,
          method  = 'circle',
          tl.cex  = .75,
          tl.col  = 'gray50',
          cl.pos  = 'n',
          addgrid = NA
        ) 
      } else {
        cols = apply(Z, c(1, 2), function(topicvalue)
          ifelse(topicvalue == 1, '#053061', '#67001F'))
        plot(
          1:nrow(wordmat),
          1:ncol(wordmat),
          type = "n",
          axes = FALSE,
          xlab = '',
          ylab = ''
        )
        text(col(wordmat),
             rev(row(wordmat)),
             wordmat,
             col = cols,
             cex = .75)
      }
    }
      
    
    # for every document and every word in the document
    for (i in 1:d) {
      for (j in 1:length(wordmat[i,])) {
        word.id = which(vocab == wordmat[i, j])
        topic.old = Z[i, j]
        
        # Decrement counts before computing equation (3) in paper noted above
        C_dt[i, topic.old] = C_dt[i, topic.old] - 1
        C_wt[word.id, topic.old] = C_wt[word.id, topic.old] - 1
        
        # Calculate equation (3) for each topic
        vals = prop.table(C_wt + beta, 2)[word.id,] * prop.table(C_dt[i,] + alpha)
        
        # Sample the new topic from the results for (3);         
        # note, sample function does not require you to have the probs sum to 1
        # explicitly, i.e.  prob=c(1,1,1) is the same as prob = c(1/3, 1/3, 1/3)
        Z.new = sample(1:K, 1, prob = vals)
        
        # Set the new topic and update counts
        Z[i,j] = Z.new
        C_dt[i, Z.new] = C_dt[i, Z.new] + 1
        C_wt[word.id, Z.new] = C_wt[word.id, Z.new] + 1
      }
    }
    
    theta = prop.table(C_dt + alpha,1)           # doc topic distribution
    phi   = prop.table(C_wt + beta,2)            # word topic distribution
    
    # save simulations
    if (s %in% saveSim){
      thetaList[[paste(s)]] = theta
      phiList[[paste(s)]]   = phi
    }
  }
  
  layout(1) # reset plot window
  
  
  results = list(
    theta = theta,
    phi   = phi,
    thetaSims = abind::abind(thetaList, along = 3),  # abind creates arrays from the lists
    phiSims   = abind::abind(phiList, along = 3)
  )
  
  results
}
```



## Estimation

We use the values of alpha and beta as suggested in paper.  The following will result in 500 posterior sample points, so bump if you want more than that.  These settings take around 20 seconds.

```{r tm-est}
alpha = K/50
beta  = .01

fit_tm = topic_model(
  alpha  = alpha,
  beta   = beta,
  nsim   = 1000,
  warmup = 500,
  thin   = 1,
  verbose = FALSE,
  plot    = FALSE
)
```



## Comparison

First we can explore the results, in particular topic assignment probabilities and term topic probabilities.

```{r tm-inspect}
str(fit_tm, 1)
purrr::map(fit_tm[1:2], round, 2)
```

Now we can compare our result to reported paper estimates for phi and <span class="pack" style = "">topicmodels</span> package (note delta is the beta above; the vignette actually references Steyvers & Griffiths paper).

```{r tm-LDA}
library(topicmodels)

fit_lda = LDA(
  dtm,
  k = 2,
  method   = 'Gibbs',
  control  = list(
    alpha  = alpha,
    delta  = .01,
    iter   = 5500,
    burnin = 500,
    thin   = 10,
    initialize = 'random'
  )
)

```

Start with $\phi$ estimates. Note that the labels of what is topic 1 vs. topic 2 is arbitrary.


```{r tm-compare-phi, echo=FALSE}
phi_estimates = data.frame(
  fit     = round(fit_tm$phi, 2),
  paper   = cbind(c(0, 0, .39, .32, .29), c(.25, .4, .35, 0, 0)),
  ldapack = round(t(posterior(fit_lda)$terms), 2)
) 

# make arbitrary labels consistent?
# if (sum(abs(phi_estimates$fit.1 - phi_estimates$paper.1)) > .5) {
#   phi_estimates = phi_estimates %>% 
#   mutate(
#     fit.1 = phi_estimates$fit.2,
#     fit.2 = phi_estimates$fit.1
#          )
# }

kable_df(phi_estimates)
```

Interval estimates for term probabilities for each topic.


```{r tm-phi-interval}
library(coda)

phi_estimates_topic_1 = as.mcmc(t(fit_tm$phiSims[, 1, ]))
phi_estimates_topic_2 = as.mcmc(t(fit_tm$phiSims[, 2, ]))

# summary(phi_estimates_topic_1)
```


```{r tm-phi-interval-show, echo=FALSE}
map_df(list(
  `Topic 1` = round(phi_estimates_topic_1, 2),
  `Topic 2` = round(phi_estimates_topic_2, 2)
), function(x)
  data.frame(HPDinterval(x)), .id = 'Topic') %>%
  rownames_to_column() %>% 
  kable_df()

```

Symmetrized or mean [Kullback-Liebler divergence](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence) for topic (dis)similarity.

```{r tm-kldiv-topic}
KLdivs = .5 * sum(
  apply(
    fit_tm$phi, 
    1, function(row)  row[1] * log2(row[1] / row[2]) + row[2] * log2(row[2] / row[1])
    )
  )

KLdivs
```

We can compare with various other packages for KL-divergence.  

```{r tm-kldiv-topic-compare, eval = FALSE}
mean(c(
  entropy::KL.empirical(fit_tm$phi[,1], fit_tm$phi[,2], unit = 'log2'),
  entropy::KL.empirical(fit_tm$phi[,2], fit_tm$phi[,1], unit = 'log2')
  ))
```

Symmetrized or mean Kullback-Liebler divergence for document (dis)similarity.

```{r tm-kldiv-doc}
# example
docs2compare = c(1, 16)

.5 * sum(apply(fit_tm$theta[docs2compare, ], 2,
               function(topicprob)
                 topicprob[1] * log(topicprob[1] / topicprob[2]) +
                 topicprob[2] * log(topicprob[2] / topicprob[1])))
```


Let's create a function to work on theta to produce Kullback-Liebler or Jensen-Shannon divergence.

```{r tm-kldiv-doc-func}
divergence = function(input, method = 'KL') {
  if (method == 'KL') {
    .5 * sum(apply(fit_tm$theta[input, ], 2,
                   function(topicprob)
                     topicprob[1] * log2(topicprob[1] / topicprob[2]) +
                     topicprob[2] * log2(topicprob[2] / topicprob[1])))
  } 
  else {
    .5 * sum(apply(fit_tm$theta[input, ], 2,
                   function(topicprob)
                     topicprob[1] * log2(topicprob[1] / mean(topicprob)) +
                     topicprob[2] * log2(topicprob[2] / mean(topicprob))))
  }
}
```

Now apply to all pairs of documents.

```{r tm-kldiv-doc-all}
# Now do for all
docpairs = combn(1:d, 2)
KLdivs = apply(docpairs, 2, divergence)
mat0 = matrix(0, d, d)
mat0[lower.tri(mat0)] = KLdivs

KLdivs = mat0 + t(mat0)
dimnames(KLdivs) = list(row.names(dtm))
```

We can visualizes as follows, red implies documents are more dissimilar.

```{r tm-kldiv-doc-all-vis, echo = FALSE}
heatmaply::heatmaply_cor(
  KLdivs,
  Rowv = FALSE,
  Colv = FALSE,
  limits = NULL,
  main = 'Kullback-Liebler Divergence'
)
```



Likewise, we can visualize [Jensen-Shannon divergence](https://en.wikipedia.org/wiki/Jensen%E2%80%93Shannon_divergence).

```{r tm-js-div-doc-all}
js_divergence = apply(docpairs, 2, divergence, method = 'JS')
```

```{r tm-js-div-doc-all-vis, echo = FALSE}
mat0 = matrix(0, d, d)
mat0[lower.tri(mat0)] = js_divergence
js_divergence = mat0 + t(mat0)
dimnames(js_divergence) = list(row.names(dtm))

heatmaply::heatmaply_cor(
  js_divergence,
  Rowv = FALSE,
  Colv = FALSE,
  limits = NULL,
  main = 'Jensen-Shannon Divergence'
)
```


Here are examples of diagnostics with the word topic probability estimates.  These use the <span class="" style = "">bayesplot</span> package.

```{r tm-bayes-diagnostics}
library(bayesplot)

mcmc_combo(phi_estimates_topic_1)
mcmc_acf(phi_estimates_topic_1)
```

## Source

Original code available at:
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/Bayesian/topicModelgibbs.R