# Cox Survival

Some simple demonstrations of a standard Cox, Cox with time-varying covariates and a stratified Cox.

## Standard Proportional Hazards

### Data Setup

```{r cox-setup}
set.seed(12)

dur = 1:10
kittyblarg = rnorm(10)                                 # something happened to kitty!
kittyhappy = rep(0:1, times = 5)                       # is kitty happy?
kittydied  = sample(0:1, 10, replace = TRUE)           # kitty died! oh no!
d = data.frame(kittyblarg, kittyhappy, dur, kittydied)

# Inspect
d
```



### Function

Create a the likelihood function to feed to <span class="" style = "">optim</span>.

```{r cox-ll}
pl <- function(pars, preds, died, t) {
  # Arguments- 
  # pars: coefficients of interest
  # preds: predictor matrix
  # died: death
  # t: time
  
  b = pars
  X = as.matrix(preds[order(t), ])
  died2 = died[order(t)]
  
  LP = X%*%b                # Linear predictor
  
  # initialize log likelihood due to looping, not necessary
  ll = numeric(nrow(X))     
  rows = 1:nrow(preds)
  
  for (i in rows){
    riskset = ifelse(rows < i, FALSE, TRUE)                  # identify risk set
    ll[i] = died2[i]*(LP[i] - log(sum(exp(LP[riskset]))) )   # log likelihood
  }
  
  -sum(ll)
}
```




### Estimation

Estimate with <span class="func" style = "">optim</span>.

```{r cox-ph-est}
initial_values = c(0, 0)

out = optim(
  par   = initial_values,
  fn    = pl,
  preds = d[, c('kittyblarg', 'kittyhappy')],
  died  = d[, 'kittydied'],
  t     = dur,
  method  = "BFGS",
  hessian = T
)

out
```



### Comparison


Extract results.

```{r cox-ph-extract}
B  = out$par
se = sqrt(diag(solve(out$hessian)))
Z  = B/se

# create a summary table

result_tbl = data.frame(
  B, 
  exp = exp(B), 
  se, 
  Z,
  p = ifelse(Z > 0, pnorm(Z, lower = FALSE)*2, pnorm(Z, lower = TRUE)*2) 
)
```



Compare to <span class="pack" style = "">survival</span> package.

```{r coxph-compare-surv}
library(survival)

cox_model = coxph(Surv(dur, kittydied) ~ kittyblarg + kittyhappy)

summary(cox_model)$coefficients
result_tbl
```





## Time-varying coefficients

Note that technically nothing new is going on here relative to the previous model.  See the vignette for the <span class="pack" style = "">survival</span> package for further details.


### Data Setup

In the following we'll first create some noisy time points.


```{r coxtv-setup}
set.seed(123)

t1 = rep(NA, 20)
t2 = rep(NA, 20)

t1[seq(1, 20, by = 2)] = 1:10
t2[seq(1, 20, by = 2)] = t1[seq(1, 20, by = 2)] + 
  sample(1:5, 10, replace = TRUE) + 
  abs(rnorm(10))

t1[seq(2, 20, by = 2)] = t2[seq(1, 20, by = 2)]
t2[seq(2, 20, by = 2)] = t1[seq(2, 20, by = 2)] + sample(1:5) + abs(rnorm(10))

kitty = rep(1:10, e = 2)
kittyblarg = t2 + rnorm(20, sd = 5)
kittyhappy = rep(0:1, times = 5, e = 2)
die = 0:1
cens = c(0, 0)
kittydied = ifelse(runif(20)>=.5, die, cens)

d = data.frame(kitty, kittyblarg, kittyhappy, 
               t1, t2, kittydied)

# Inspect the Surv object if desired
# Surv(t1,t2, kittydied)

# Inspect the data
d
```





### Function

```{r pltv-func}
pl_tv <- function(pars, preds, died, t1, t2, data) {
  # Same arguments as before though will take a data object
  # plus variable names via string input. Also requires beginning
  # and end time point (t1, t2)
  dat = data[,c(preds, died, t1, t2)]
  dat = dat[order(dat$t2), ]
  b   = pars
  X   = as.matrix(dat[, preds])
  died2 = dat[, died]
  
  # linear predictor
  LP = X%*%b
  
  # log likelihood
  ll = numeric(nrow(X))
  rows = 1:nrow(dat)
  
  for (i in rows){
    st_i = dat$t2[i]
    
    # if they have already died/censored (row < i) or if the initial time is
    # greater than current end time (t1 > st_i),  they are not in the risk set,
    # else they are.
    riskset = ifelse(rows < i | dat$t1 > st_i, FALSE, TRUE)     
    
    ll[i] = died2[i]*(LP[i] - log(sum(exp(LP[riskset]))) )        
  }     
  
  -sum(ll)
}

```


### Estimation

Estimate with <span class="func" style = "">optim</span>.

```{r coxtv-est}
initial_values = c(0, 0)

out = optim(
  par   = initial_values,
  fn    = pl_tv,
  preds = c('kittyblarg', 'kittyhappy'),
  died  = 'kittydied',
  data  = d,
  t1    = 't1',
  t2    = 't2',
  method  = "BFGS",
  hessian = TRUE
)

# out
```

### Comparison


Extract results.

```{r coxtv-extract}
B  = out$par
se = sqrt(diag(solve(out$hessian)))
Z  = B/se

result_tbl = data.frame(
  B, 
  exp = exp(B), 
  se, 
  Z, 
  p = ifelse(Z > 0, pnorm(Z, lower = FALSE) * 2, pnorm(Z, lower = TRUE) * 2)
)
```



Compare to survival package.

```{r coxtv-compare}
cox_model_tv = coxph(
  Surv(t1, t2, kittydied) ~ kittyblarg + kittyhappy,
  method = 'breslow',
  control = coxph.control(iter.max = 1000)
)
# cox_model_tv; cox_model_tv$loglik[2]

summary(cox_model_tv)$coefficients

result_tbl
```




## Stratified Cox Model


### Data Setup

```{r coxstrat-setup}
data(ovarian, package = 'survival') 
```

### Function

Requires `pl` function above though one could extend to `pl_tv`.

```{r pl-strat-func}
pl_strat <- function(pars, preds, died, t, strata) {
  strat = as.factor(strata)
  d = data.frame(preds, died, t, strat)
  dlist = split(d, strata)
  
  neglls = map_dbl(
    dlist,
    function(x)
      pl(
        pars  = pars,
        preds = x[, colnames(preds)],
        died  = x$died,
        t     = x$t
      )
  )
  
  sum(neglls)
}
```



### Estimation

Estimate with <span class="func" style = "">optim</span>.

```{r coxstrat-est}
initial_values = c(0, 0)

out = optim(
  par   = initial_values,
  fn    = pl_strat,
  preds = ovarian[, c('age', 'ecog.ps')],
  died  = ovarian$fustat,
  t     = ovarian$futime,
  strata  = ovarian$rx,
  method  = "BFGS",
  hessian = TRUE
)

# out
```

### Comparison

```{r coxstrat-compare}
B  = out$par
se = sqrt(diag(solve(out$hessian)))
Z  = B/se

results_tbl = data.frame(
  B,
  exp = exp(B),
  se,
  Z,
  p = ifelse(Z > 0, pnorm(Z, lower = FALSE) * 2, pnorm(Z, lower = TRUE)*2)
)
  

cox_strata_model = coxph(
  Surv(futime, fustat) ~ age + ecog.ps + strata(rx), 
  data = ovarian
  )
# cox_strata_model; cox_strata_model$loglik[2]

summary(cox_model_tv)$coefficients

result_tbl
```


## Source

Original code available at
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/survivalCox.R
