# Extreme Learning Machine

A very simple implementation of an extreme learning machine for regression. See <span class="pack" style = "">elmNN</span> and <span class="pack" style = "">ELMR</span> for some R package implementations. I add comparison to generalized additive models (elm/neural networks and GAMs are adaptive basis function models). 


- http://www.extreme-learning-machines.org
- G.B. Huang, Q.Y. Zhu and C.K. Siew, Extreme Learning Machine: Theory and Applications.


## Data Setup


One variable, complex function.

```{r elm-setup1}
library(tidyverse)
library(mgcv)

set.seed(123)

n = 5000
x = runif(n)
# x = rnorm(n)
mu = sin(2*(4*x-2)) + 2* exp(-(16^2) * ((x-.5)^2))
y  = rnorm(n, mu, .3)

d = data.frame(x, y) 

qplot(x, y, color = I('#ff55001A'))
```


Motorcycle accident data.

```{r elm-setup2}
data('mcycle', package = 'MASS')

times = matrix(mcycle$times, ncol = 1)
accel = mcycle$accel
```



## Function

```{r elm-func}
elm <- function(X, y, n_hidden = NULL, active_fun = tanh) {
  # X: an N observations x p features matrix
  # y: the target
  # n_hidden: the number of hidden nodes
  # active_fun: activation function
  
  pp1 = ncol(X) + 1
  w0  = matrix(rnorm(pp1*n_hidden), pp1, n_hidden)       # random weights
  h   = active_fun(cbind(1, scale(X)) %*% w0)            # compute hidden layer
  B   = MASS::ginv(h) %*% y                              # find weights for hidden layer
  
  fit = h %*% B                                          # fitted values
  
  list(
    fit  = fit,
    loss = crossprod(y - fit),
    B    = B,
    w0   = w0
  )
}
```




## Estimation

```{r elm-est1}
X_mat = as.matrix(x, ncol = 1)

fit_elm = elm(X_mat, y, n_hidden = 100)
str(fit_elm)

ggplot(aes(x, y), data = d) +
  geom_point(color = '#ff55001A') + 
  geom_line(aes(y = fit_elm$fit), color = '#00aaff')

cor(fit_elm$fit[,1], y)^2
```


```{r  elm-est2}
fit_elm_mcycle = elm(times, accel, n_hidden = 100)
cor(fit_elm_mcycle$fit[,1], accel)^2
```




## Comparison

We'll compare to a generalized additive model with gaussian process approximation.

```{r elm-compare1}
fit_gam = gam(y ~ s(x, bs = 'gp', k = 20), data = d)
summary(fit_gam)$r.sq


d %>%
  mutate(fit_elm = fit_elm$fit,
         fit_gam = fitted(fit_gam)) %>%
  ggplot() +
  geom_point(aes(x, y), color = '#ff55001A') +
  geom_line(aes(x, y = fit_elm), color = '#1e90ff') +
  geom_line(aes(x, y = fit_gam), color = '#990024')
```


```{r elm-compare2}
fit_gam_mcycle = gam(accel ~ s(times), data = mcycle)
summary(fit_gam_mcycle)$r.sq

mcycle %>% 
  ggplot(aes(times, accel)) +
  geom_point(color = '#ff55001A') +
  geom_line(aes(y = fit_elm_mcycle$fit), color = '#1e90ff') +
  geom_line(aes(y = fitted(fit_gam_mcycle)), color = '#990024')
```



## Supplemental Example

Yet another example with additional covariates.

```{r elm-setup-3}
d = gamSim(eg = 7, n = 10000)
X = as.matrix(d[, 2:5])
y = d[, 1]

n_nodes = c(10, 25, 100, 250, 500, 1000)
```

The following estimation over multiple models will take several seconds.

```{r elm-est3}
fit_elm = purrr::map(n_nodes, function(n) elm(X, y, n_hidden = n))
```

Now find the best fitting model.

```{r elm-best}
# estimate 
best_loss = which.min(map_dbl(fit_elm, function(x) x$loss))
fit_best  = fit_elm[[best_loss]]
```

A quick check of the fit.

```{r elm-fit}
# str(fit_best)
# qplot(fit_best$fit[, 1], y, alpha = .2)
cor(fit_best$fit[, 1], y)^2
```


And compare again to <span class="pack" style = "">mgcv</span>.  In this case, we're comparing fit on test data of the same form.

```{r elm-compare3}
fit_gam = gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = d)
gam.check(fit_gam)
summary(fit_gam)$r.sq


test_data0 = gamSim(eg = 7)  # default n = 400
test_data  = cbind(1, scale(test_data0[, 2:5]))

 # remember to use your specific activation function here
elm_prediction = tanh(test_data %*% fit_best$w0) %*% fit_best$B         
gam_prediction = predict(fit_gam, newdata = test_data0)

cor(data.frame(elm_prediction, gam_prediction), test_data0$y)^2
```



## Source

Original code available at:
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/elm.R
