## Mixture Model

The following code is based on algorithms noted in Murphy, 2012 Probabilistic Machine Learning, specifically, Chapter 11, section 4.



### Data Setup

This example uses Old Faithful geyser eruptions.  This is only a univariate
mixture for either eruption time or wait time. The [next example][Multivariate
Mixture Model] will be doing both variables, i.e. multivariate normal.  'Geyser'
is supposedly more accurate, though seems to have arbitrarily assigned some
duration values.  See [this
source](http://www.geyserstudy.org/geyser.aspx?pGeyserNo=OLDFAITHFUL) also, but
that only has intervals. Some July 1995 data is available.

```{r old-faithful}
library(tidyverse)

# faithful data set is in base R
data(faithful)

head(faithful)

eruptions  = as.matrix(faithful[, 1, drop = FALSE])
wait_times = as.matrix(faithful[, 2, drop = FALSE])
```

### Function

The fitting function.

```{r em_mixture}
em_mixture <- function(
  params,
  X,
  clusters = 2,
  tol = .00001,
  maxits  = 100,
  showits = TRUE
) {
  
  # Arguments are starting parameters (means, covariances, cluster probability),
  # data, number of clusters desired, tolerance, maximum iterations, and whether
  # to show iterations
  
  # Starting points
  N     = nrow(X)
  nams  = names(params)
  mu    = params$mu
  var   = params$var
  probs = params$probs
  
  # Other initializations
  # initialize cluster 'responsibilities', i.e. probability of cluster
  # membership for each observation i
  ri = matrix(0, ncol = clusters, nrow = N) 
  it = 0
  converged = FALSE
  
  if (showits)                                  # Show iterations
    cat(paste("Iterations of EM:", "\n"))
  
  while ((!converged) & (it < maxits)) { 
    probsOld = probs
    muOld = mu
    varOld = var
    riOld = ri
    
    # E
    # Compute responsibilities
    for (k in 1:clusters){
      ri[, k] = probs[k] * dnorm(X, mu[k], sd = sqrt(var[k]), log = FALSE)
    }
    
    ri = ri/rowSums(ri)
    
    # M
    rk = colSums(ri)           # rk is the weighted average cluster membership size
    probs = rk/N
    mu = (t(X) %*% ri) / rk    
    var = (t(X^2) %*% ri) / rk - mu^2
    
    # could do mu and var via log likelihood here, but this is more straightforward

    parmlistold     = rbind(probsOld, muOld, varOld)
    parmlistcurrent = rbind(probs, mu, var)
    
    it = it + 1
    
    # if showits true, & it =1 or divisible by 5 print message
    if (showits & it == 1 | it%%5 == 0)        
      cat(paste(format(it), "...", "\n", sep = ""))
    
    converged = max(abs(parmlistold - parmlistcurrent)) <= tol
  }
  
  clust = which(round(ri) == 1, arr.ind = TRUE) # create cluster membership
  clust = clust[order(clust[, 1]), 2]           # order according to row rather than cluster
  
  out = list(
    probs   = probs,
    mu      = mu,
    var     = var,
    resp    = ri,
    cluster = clust
  )
  
  out
} 
```

### Estimation

Starting parameters require mean, variance and class probability. Note that starting values for mean must be within the data range or it will break.  

```{r gaussEM-starts}
start_values_1 = list(mu = c(2, 5),
                      var = c(1, 1),
                      probs = c(.5, .5))

start_values_2 = list(mu = c(50, 90),
                      var = c(1, 15),
                      probs = c(.5, .5)) 

```


```{r gaussEM-est}
mix_erupt   = em_mixture(start_values_1, X = eruptions,  tol = 1e-8)
mix_waiting = em_mixture(start_values_2, X = wait_times, tol = 1e-8)
```



### Comparison

Compare to <span class="pack" style = "">flexmix</span> package results.

```{r gaussEM-flex}
library(flexmix)

flex_erupt = flexmix(eruptions ~ 1,
                     k = 2,
                     control = list(tolerance = 1e-8, iter.max = 100))

flex_wait = flexmix(wait_times ~ 1,
                    k = 2,
                    control = list(tolerance = 1e-8, iter.max = 100))
```


The following provides means, variances and probability of group membership. Note that the cluster label is arbitrary so cluster 1 for one model may be cluster 2 in another.

##### Eruptions

First we'll compare results on the eruptions variable.

```{r gaussEM-compare-erupt}
mean_var = rbind(mix_erupt$mu, sqrt(mix_erupt$var))
rownames(mean_var) = c('means', 'variances')
colnames(mean_var) = c('cluster 1', 'cluster 2')

mean_var_flex = parameters(flex_erupt)
rownames(mean_var_flex) = c('means', 'variances')
colnames(mean_var_flex) = c('cluster 1 flex', 'cluster 2 flex')


prob_membership      = mix_erupt$probs
prob_membership_flex = flex_erupt@size / sum(flex_erupt@size)

list(
  params = cbind(mean_var, mean_var_flex),
  clusterpobs = cbind(prob_membership, prob_membership_flex)
)
```



##### Waiting

Now we compare the result for waiting times.

```{r gaussEM-compare-wait} 
mean_var = rbind(mix_waiting$mu, sqrt(mix_waiting$var))
rownames(mean_var) = c('means', 'variances')
colnames(mean_var) = c('cluster 1', 'cluster 2')


mean_var_flex = parameters(flex_wait)
rownames(mean_var_flex) = c('means', 'variances')
colnames(mean_var_flex) = c('cluster 1 flex', 'cluster 2 flex')

prob_membership      = mix_waiting$probs
prob_membership_flex = flex_wait@size / sum(flex_wait@size)

list(
  params = cbind(mean_var, mean_var_flex),
  clusterpobs = cbind(prob_membership, prob_membership_flex)
)


```

### Visualization

Points are colored by class membership, followed by the probability of being in cluster 1.

```{r gaussEM-vis, echo=FALSE}
# qplot(x = eruptions, y = waiting, data = faithful)

faithful %>%
  mutate(cluster = factor(mix_waiting$cluster)) %>% 
  ggplot(aes(x = eruptions, y = waiting) ) +
  geom_density2d(color = 'gray92') +
  geom_point(aes(color = cluster)) +
  scico::scale_color_scico_d(begin = .25, end = .75, alpha = .5) +
  guides(color = guide_legend('Cluster'))


faithful %>%
  mutate(prob_clus_1 = mix_waiting$resp[, 1]) %>%
  ggplot(aes(x = eruptions, y = waiting)) +
  geom_density2d(color = 'gray92') +
  geom_point(aes(color = prob_clus_1)) +
  scico::scale_color_scico(palette = 'bilbao', begin = .25) +
  guides(color = guide_legend('Prob. Cluster 1'))
```


### Supplemental Example

This uses the <span class="pack" style = "">MASS</span> version (reversed columns). These don't look even remotely the same data on initial inspection- `geyser` is even more rounded and of opposite conclusion.  Turns out geyser is offset by 1, such that duration 1 should be coupled with waiting 2 and on down. Still the rounding at 2 and 4 (and whatever division was done on duration) makes this fairly poor data.

I've cleaned this up a little bit in case someone wants to play with it for additional practice, but it's not evaluated.

```{r mass-demo, eval=FALSE}
data(geyser, package = 'MASS')

geyser = data.frame(duration = geyser$duration[-299], waiting = geyser$waiting[-1])

# compare to faithful
library(patchwork)

qplot(data = faithful, x = eruptions, y = waiting, alpha = I(.25)) /
  qplot(data = geyser, x = duration,  y = waiting, alpha = I(.25))

X3 = matrix(geyser[,1]) 
X4 = matrix(geyser[,2])


# MASS version
test3 = em_mixture(start_values_1, X = X3, tol = 1e-8)
test4 = em_mixture(start_values_2, X = X4, tol = 1e-8)

flexmod3 = flexmix(X3 ~ 1,
                   k = 2,
                   control = list(tolerance = 1e-8, iter.max = 100))
flexmod4 = flexmix(X4 ~ 1,
                   k = 2,
                   control = list(tolerance = 1e-8, iter.max = 100))

# note variability differences compared to faithful dataset
# Eruptions/Duration
mean_var = rbind(test3$mu, sqrt(test3$var))
rownames(mean_var) = c('means', 'variances')

mean_var_flex = parameters(flexmod3)
rownames(mean_var_flex) = c('means', 'variances')

prob_membership = test3$probs
prob_membership_flex = flexmod3@size / sum(flexmod3@size)

list(
  params = cbind(mean_var, mean_var_flex),
  clusterpobs = cbind(prob_membership, prob_membership_flex)
)

# Waiting
mean_var = rbind(test4$mu, sqrt(test4$var))
rownames(mean_var) = c('means', 'variances')

mean_var_flex = parameters(flexmod4)
rownames(mean_var_flex) = c('means', 'variances')

prob_membership = test4$probs
prob_membership_flex = flexmod4@size / sum(flexmod4@size)

list(
  params = cbind(mean_var, mean_var_flex),
  clusterpobs = cbind(prob_membership, prob_membership_flex)
)

# Some plots
library(ggplot2)

qplot(x = eruptions, y = waiting, data = faithful)

ggplot(aes(x = eruptions, y = waiting), data = faithful) +
  geom_point(aes(color = factor(mix_waiting$cluster)))

ggplot(aes(x = eruptions, y = waiting), data = faithful) +
  geom_point(aes(color = mix_waiting$resp[, 1]))
```


### Source

Original code available at
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/EM%20Examples/EM%20Mixture.R
