## Probabilistic PCA

The following is an EM algorithm for probabilistic principal components
analysis. Based on Tipping and Bishop, 1999, and also Murphy 2012 Probabilistic
ML, with some code snippets inspired by the <span class="func" style = "">ppca</span> function used  below. See also ModelFitting/EM Examples/EM for pca.R


### Data Setup

`state.x77` is from base R, which includes various state demographics.  We will first standardize the data.

```{r em-ppca-setup}
X = scale(state.x77)
```

### Function

```{r em-ppca-func}
em_ppca = function(
  X,
  nComp = 2,
  tol   = .00001,
  maxits  = 100,
  showits = TRUE
) {
  
  # Arguments 
  # X: numeric data
  # nComp: number of components
  # tol = tolerance level
  # maxits: maximum iterations
  # showits: show iterations
  
  # require(pracma) 

  tr = function(x) sum(diag(x), na.rm = TRUE)
  
  
  # starting points and other initializations
  N = nrow(X)
  D = ncol(X)
  L = nComp

  S = (1/N) * t(X)%*%X
  
  evals = eigen(S)$values
  evecs = eigen(S)$vectors

  V = evecs[,1:L]
  Lambda = diag(evals[1:L])
  
  # latent variables
  Z = t(replicate(L, rnorm(N))) 
  
  # variance; average variance associated with discarded dimensions
  sigma2 = 1/(D - L) * sum(evals[(L+1):D])
  
  # loadings; this and sigma2 starting points will be near final estimate
  W = V %*% chol(Lambda - sigma2 * diag(L)) %*% diag(L)  

  it = 0
  converged = FALSE
  ll = 0
  
  # Show iterations
  if (showits)                                                    
    cat(paste("Iterations of EM:", "\n"))
  
  while ((!converged) & (it < maxits)) {                           
    # create 'old' values for comparison
    if(exists('W_new')){
      W_old = W_new
    }
    else {
      W_old = W
    }
    
    ll_old = ll
    
    Psi = sigma2*diag(L)
    M   = t(W_old) %*% W_old + Psi
    
    # E and M 
    W_new  = S %*% W_old %*% solve( Psi + solve(M) %*% t(W_old) %*% S %*% W_old )   
    sigma2 = 1/D * tr(S - S %*% W_old %*% solve(M) %*% t(W_new))

    Z  = solve(M) %*% t(W_new) %*% t(X)
    ZZ = sigma2*solve(M) + Z%*%t(Z)
    
    # log likelihood as in paper
    # ll = .5*sigma2*D + .5*tr(ZZ) + .5*sigma2 * X%*%t(X) -
    #      1/sigma2 * t(Z)%*%t(W_new)%*%t(X) + .5*sigma2 * tr(t(W_new)%*%W_new%*%ZZ)
    # ll = -sum(ll)
    
    # more straightforward
    ll = dnorm(X, mean = t(W_new %*% Z), sd = sqrt(sigma2), log = TRUE)
    ll = -sum(ll)

    it = it + 1
    
    # if showits, show first and every 5th iteration
    if (showits & (it == 1 | it%%5 == 0))                          
      cat(paste(format(it), "...", "\n", sep = ""))
    
    converged = max(abs(ll_old-ll)) <= tol
  }
  
  W     = pracma::orth(W_new) # for orthonormal basis of W; pcaMethods package has also
  evs   = eigen(cov(X %*% W))
  evecs = evs$vectors
  
  W = W %*% evecs
  Z = X %*% W
  Xrecon   = Z %*% t(W)
  reconerr = sum((Xrecon - X)^2)
  
  if (showits)                                                     # Show last iteration
    cat(paste0(format(it), "...", "\n"))
  
  list(
    scores   = Z,
    loadings = W,
    Xrecon   = Xrecon,
    reconerr = reconerr,
    ll       = ll,
    sigma2   = sigma2
  )
}
```








### Estimation

```{r em-ppca-est}
results_ppca = em_ppca(
  X = X,
  nComp = 2,
  tol = 1e-12,
  maxit = 100
)

str(results_ppca)
```

### Comparison

Extract reconstructed values and loadings for comparison.

```{r em-ppca-extract}
Xrecon      = results_ppca$Xrecon
loadings_em = results_ppca$loadings
scores_em   = results_ppca$scores
```


Compare to standard pca on full data set if desired.

```{r em-ppca-compare-pca}
standard_pca =  princomp(scale(state.x77))

scores_standard_pca   = standard_pca$scores[,1:2]
loadings_standard_pca = standard_pca$loadings[,1:2]
Xrecon_standard_pca   = scores_standard_pca%*%t(loadings_standard_pca)
```




Compare results to output from <span class="pack" style = "">pcaMethods</span>, which also has probabilistic PCA (demonstrated next). Note that the signs for loadings/scores may be different


```{r em-ppca-compare}
library(pcaMethods)

results_pcam  = pca(
  X,
  nPcs = 2,
  threshold = 1e-8,
  method = 'ppca',
  scale  = 'none',
  center = FALSE
)

loadings_pcam = loadings(results_pcam)
scores_pcam   = scores(results_pcam)
```


Compare loadings and scores.

```{r em-ppca-compare-loadings}
round(cbind(loadings_pcam, loadings_em, loadings_standard_pca), 3)
sum((abs(loadings_pcam) - abs(loadings_em)) ^ 2)
round(cbind(abs(scores_pcam), abs(scores_em)), 2)
```


Compare reconstructed data sets.

```{r em-ppca-compare-recon}
Xrecon_pcam = scores_pcam %*% t(loadings_pcam)

mean((Xrecon_pcam - X)^2)  
mean(abs(Xrecon_pcam - Xrecon))

mean(abs(Xrecon_pcam - Xrecon))
```



### Visualize

```{r em-ppca-vis-recon}
GGally::ggpairs(data.frame(
  data    = X[, 1],
  custom  = Xrecon[, 1],
  pcaMeth = Xrecon_pcam[, 1]
))

GGally::ggpairs(data.frame(
  data    = X[, 2],
  custom  = Xrecon[, 2],
  pcaMeth = Xrecon_pcam[, 2]
))
```

```{r em-ppca-vis-scores}
qplot(Xrecon[, 1], Xrecon_pcam[, 1])

GGally::ggpairs(data.frame(scores_em, scores_pcam) )
```





### Missing Data Example

A slightly revised approach can be taken in the case of missing value.


#### Data Setup

```{r em-ppca-miss-setup}
# create some missing values
set.seed(123)

X_miss = X
NAindex = sample(length(X), 20)
X_miss[NAindex] = NA
```


#### Function

```{r em-ppca-miss-func}
em_ppca_miss = function(X, nComp=2, tol=.00001, maxits=100, showits=T){
  # Arguments 
  # X: numeric data
  # nComp: number of components
  # tol = tolerance level
  # maxits: maximum iterations
  # showits: show iterations
  # require(pracma) # for orthonormal basis of W; pcaMethods package has also
  
  tr = function(x) sum(diag(x), na.rm = TRUE)
  
  # starting points and other initializations
  X_orig = X
  X = X
  N = nrow(X_orig)
  D = ncol(X_orig)
  L = nComp
  NAs = is.na(X_orig)

  X[NAs] = 0
  S = (1/N) * t(X)%*%X
  
  evals = eigen(S)$values
  evecs = eigen(S)$vectors

  V = evecs[,1:L]
  Lambda = diag(evals[1:L])
  
  # latent variables
  Z = t(replicate(L, rnorm(N))) 
  
  # variance; average variance associated with discarded dimensions
  sigma2 = 1/(D-L) * sum(evals[(L+1):D])
  
  # loadings
  W = V %*% chol(Lambda-sigma2*diag(L)) %*% diag(L)           

  it = 0
  converged = FALSE
  ll = 0
  
   # Show iterations
  if (showits)
    cat(paste("Iterations of EM:", "\n"))
  
  while ((!converged) & (it < maxits)) {                  
    if(exists('W_new')){
      W_old = W_new
    }
    else {
      W_old = W
    }
    
    ll_old = ll
    
    # deal with missingness via projection
    proj  = t(W_old%*%Z)
    X_new = X_orig
    X_new[NAs] = proj[NAs]
    X = X_new
    
    Psi = sigma2*diag(L)
    M   = t(W_old) %*% W_old + Psi
    
    # E and M
    W_new  = S %*% W_old %*% solve( Psi + solve(M)%*%t(W_old)%*%S%*%W_old )
    sigma2 = 1/D * tr(S - S%*%W_old%*%solve(M)%*%t(W_new))

    Z = solve(M)%*%t(W_new)%*%t(X)
    
  
    # log likelihood as in paper
    # ZZ = sigma2*solve(M) + Z%*%t(Z)
    # ll = .5*sigma2*D + .5*tr(ZZ) + .5*sigma2 * X%*%t(X) -
    #      1/sigma2 * t(Z)%*%t(W_new)%*%t(X) + .5*sigma2 * tr(t(W_new)%*%W_new%*%ZZ)
    # ll = -sum(ll)
    
    # more straightforward
    ll = dnorm(X, mean = t(W_new %*% Z), sd = sqrt(sigma2), log = TRUE)
    ll = -sum(ll)

    it = it + 1
    
    # if showits, show first and every 5th iteration
    if (showits & (it == 1 | it%%5 == 0))                          
      cat(paste(format(it), "...", "\n", sep = ""))
    
    converged = max(abs(ll_old-ll)) <= tol
  }
  
  W     = pracma::orth(W_new)   # for orthonormal basis of W
  evs   = eigen(cov(X %*% W))
  evecs = evs$vectors
  
  W = W %*% evecs
  Z = X %*% W
  Xrecon   = Z %*% t(W)
  reconerr = sum((Xrecon-X)^2)
  
  if (showits)                                                     # Show last iteration
    cat(paste0(format(it), "...", "\n"))
  
  list(
    scores   = Z,
    loadings = W,
    Xrecon   = Xrecon,
    reconerr = reconerr,
    ll       = ll,
    sigma2   = sigma2
  )
}
```


#### Estimation

Run the PCA.

```{r em-ppca-miss-est}
results_ppca_miss = em_ppca_miss(
  X = X_miss,
  nComp = 2,
  tol   = 1e-8,
  maxit = 100
)

str(results_ppca_miss)
```



#### Comparison

Extract reconstructed values and loadings for comparison.

```{r em-ppca-miss-extract}
Xrecon      = results_ppca_miss$Xrecon
loadings_em = results_ppca_miss$loadings
scores_em   = results_ppca_miss$scores
```


Compare to standard pca on full data set if desired.

```{r em-ppca-miss-compare-pca, eval=FALSE}
standard_pca =  princomp(scale(state.x77))

scores_standard_pca   = standard_pca$scores[,1:2]
loadings_standard_pca = standard_pca$loadings[,1:2]
Xrecon_standard_pca   = scores_standard_pca%*%t(loadings_standard_pca)
```




Compare results to output from <span class="pack" style = "">pcaMethods</span>, which also has probabilistic PCA (demonstrated next). Note that the signs for loadings/scores may be different


```{r em-ppca-miss--compare}
library(pcaMethods)

results_pcam  = pca(
  X_miss,
  nPcs = 2,
  threshold = 1e-8,
  method = 'ppca',
  scale  = 'none',
  center = FALSE
)

loadings_pcam = loadings(results_pcam)
scores_pcam   = scores(results_pcam)
```


Compare loadings and scores.

```{r em-ppca-miss-compare-loadings}
round(cbind(loadings_pcam, loadings_em, loadings_standard_pca), 3)
sum((abs(loadings_pcam) - abs(loadings_em)) ^ 2)
round(cbind(abs(scores_pcam), abs(scores_em)), 2)
```


Compare reconstructed data sets.

```{r em-ppca-miss-compare-recon}
Xrecon_pcam = scores_pcam %*% t(loadings_pcam)

mean((Xrecon_pcam - X_miss)^2)  
mean(abs(Xrecon_pcam - Xrecon))

mean(abs(Xrecon_pcam - Xrecon))
```



#### Visualize

```{r em-ppca-miss-vis-recon}
GGally::ggpairs(data.frame(
  data    = X_miss[, 1],
  custom  = Xrecon[, 1],
  pcaMeth = Xrecon_pcam[, 1]
))

GGally::ggpairs(data.frame(
  data    = X_miss[, 2],
  custom  = Xrecon[, 2],
  pcaMeth = Xrecon_pcam[, 2]
))
```

```{r em-ppca-miss-vis-scores}
qplot(Xrecon[, 1], Xrecon_pcam[, 1])

GGally::ggpairs(data.frame(scores_em, scores_pcam) )
```


### Source

Source code found at
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/EM%20Examples/EM%20algorithm%20for%20ppca.R

Source code for the missing example found at
(https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/EM%20Examples/EM%20algorithm%20for%20ppca%20with%20missing.R