## State Space Model

The following regards chapter 11 in Statistical Modeling and Computation, the
first example for an unobserved components model.  The data regards  inflation
based on the U.S. consumer price index ($\textrm{infl} = 400*log(cpi_t/cpi_{t-1})$), from
the second quarter of 1947 to the second  quarter of 2011.  You can acquire the
data [here](http://www.maths.uq.edu.au/~kroese/statbook/Statespace/USCPI.csv) or
in [Datasets repo](https://github.com/m-clark/Datasets). Just note that it has 2
mystery columns and one mystery row presumably supplied by Excel.  You can also
get the CPI data yourself at the [Bureau of Labor
Statistics](http://www.bls.gov/cpi/) in a frustrating fashion, or in a much
easier fashion [here](https://fred.stlouisfed.org/series/CPIAUCSL).


For the following I use `n` instead of `t` or `T` because those are transpose and `TRUE`
in R. The model is basically y = τ + ϵ, with ϵ ~ N(0, σ^2^), and τ = τ<sub>n-1<sub> +
υ_n with υ ~ N(0, ω^2^).  Thus each y is associated with a latent variable that
follows a random walk over time. ω^2^ serves as a smoothing parameter, which
itself may be estimated but which is fixed in the following. See the text for
more details.


### Data Setup

```{r em-ssm-setup}
library(tidyverse)

d = read_csv(
  'https://raw.githubusercontent.com/m-clark/Datasets/master/us%20cpi/USCPI.csv',
  col_names = FALSE
)

inflation = as.matrix(d$X1)

summary(inflation)
```



### Function

EM function for a state space model.

```{r em-ssm}
em_state_space <- function(
  params,
  y,
  omega2_0,
  omega2,
  tol     = .00001,
  maxits  = 100,
  showits = FALSE
) {
  
  # Arguments 
  # params: starting parameters (variance as 'sigma2'), 
  # y: data, 
  # tol: tolerance,
  # omega2: latent variance (2_0) is a noisier starting variance
  # maxits: maximum iterations
  # showits: whether to show iterations
  
  # Not really needed here, but would be a good idea generally to take advantage
  # of sparse representation for large data
  # require(spam)  # see usage below
  
  # Starting points
  n = length(y)
  sigma2 = params$sigma2
  
  # Other initializations
  H = diag(n)
  
  for (i in 1:(ncol(H) - 1)) {
    H[i + 1, i] = -1
  }
  
  Omega2 = spam::as.spam(diag(omega2, n))
  Omega2[1, 1] = omega2_0
  
  H = spam::as.spam(H)
  HinvOmega2H = t(H) %*% spam::chol2inv(spam::chol(Omega2)) %*% H   # tau ~ N(0, HinvOmmega2H^-1)
  
  it = 0
  converged = FALSE
  
  if (showits)                                    # Show iterations
    cat(paste("Iterations of EM:", "\n"))

  while ((!converged) & (it < maxits)) { 
    sigma2Old    = sigma2[1]
    Sigma2invOld = diag(n)/sigma2Old

    K   = HinvOmega2H + Sigma2invOld              # E
    tau = solve(K, y/sigma2Old)                   # tau|y, sigma2_{n-1}, omega2 ~ N(0, K^-1)
    K_inv_tr = sum(1/eigen(K)$values)
    
    sigma2 = 1/n * (K_inv_tr + crossprod(y-tau))  # M
    
    converged = max(abs(sigma2 - sigma2Old)) <= tol
    
    it = it + 1
    
    # if showits true, & it =1 or divisible by 5 print message
    if (showits & it == 1 | it%%5 == 0)        
      cat(paste(format(it), "...", "\n", sep = ""))
  }
  
  Kfinal   = HinvOmega2H + diag(n) / sigma2[1]
  taufinal = solve(K, (y / sigma2[1]))
  
  list(sigma2 = sigma2, tau = taufinal)
}
```




### Estimation

```{r em-ssm-est}
ss_mod_1 = em_state_space(
  params = data.frame(sigma2 = var(inflation)),
  y      = inflation,
  tol    = 1e-10,
  omega2_0 = 9,
  omega2   = 1^2
)

ss_mod_.5 = em_state_space(
  params = data.frame(sigma2 = var(inflation)),
  y      = inflation,
  tol    = 1e-10,
  omega2_0 = 9,
  omega2   = .5^2
)

# more smooth
ss_mod_.1 = em_state_space(
  params = data.frame(sigma2 = var(inflation)),
  y      = inflation,
  tol    = 1e-10,
  omega2_0 = 9,
  omega2   = .1^2
)

ss_mod_1$sigma2
ss_mod_.5$sigma2
ss_mod_.1$sigma2
```




### Visualization

```{r em-ssm-vis-setup}
library(lubridate)

series = ymd(
  paste0(
    rep(1947:2014, e = 4), 
    '-', 
    c('01', '04', '07', '10'), 
    '-', 
    '01')
  )
```



The following corresponds to Fig. 11.1 in the text.  We'll also compare to generalized additive model via <span class="func" style = "">geom_smooth</span> (greenish line).  We can see the .1 model (light blue) is overly smooth.

```{r em-ssm-vis}
library(tidyverse)

data.frame(
  series = series[1:length(inflation)],
  inflation = inflation,
  Mod_1  = ss_mod_1$tau,
  Mod_.5 = ss_mod_.5$tau,
  Mod_.1 = ss_mod_.1$tau
) %>% 
  ggplot(aes(x = series, y = inflation)) +
  geom_point(color = 'gray50', alpha = .25) +
  geom_line(aes(y = Mod_1),  color = '#ff5500') +
  geom_line(aes(y = Mod_.5), color = '#9E1B34') +
  geom_line(aes(y = Mod_.1), color = '#00aaff') +
  geom_smooth(
    formula = y ~ s(x, bs = 'gp', k = 50),
    se = FALSE,
    color = '#1b9e86',
    method = 'gam',
    size = .5
  ) +
  scale_x_date(date_breaks = '10 years') +
  labs(x = '')
```


### Source

Original code found at
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/EM%20Examples/EM%20for%20state%20space%20unobserved%20components.R
