# Hamiltonian Monte Carlo


The following demonstrates Hamiltonian Monte Carlo, the technique that Stan
uses, and which is a different estimation approach than the Gibbs sampler in
BUGS/JAGS.  If you are interested in the details enough to be reading this, I
highly recommend Betancourt's [conceptual introduction to
HMC](https://arxiv.org/pdf/1701.02434.pdf).  This example is largely based on
the code in the appendix of Gelman.

- Gelman et al. 2013. Bayesian Data Analysis. 3rd ed.

## Data Setup

As with the [Metropolis Hastings chapter][Metropolis Hastings], we use some data for a standard linear regression.

```{r hmc-setup}
library(tidyverse)

# set seed for replicability
set.seed(8675309)

# create a N x k matrix of covariates
N = 250
K = 3

covariates = replicate(K, rnorm(n = N))
colnames(covariates) = c('X1', 'X2', 'X3')

# create the model matrix with intercept
X = cbind(Intercept = 1, covariates)

# create a normally distributed variable that is a function of the covariates
coefs = c(5, .2, -1.5, .9)
sigma = 2
mu = X %*% coefs
y  = rnorm(N, mu, sigma)

# same as
# y = 5 + .2*X1 - 1.5*X2 + .9*X3 + rnorm(N, mean = 0, sd = 2)

# Run lm for later comparison; but go ahead and examine now if desired
fit_lm = lm(y ~ ., data = data.frame(X[, -1]))
# summary(fit_lm)
```


## Functions

First we start with the log posterior function.

```{r hmc-func-log-post}
log_posterior <- function(X, y, th) {
  # Args
  # X: the model matrix
  # y: the target vector
  # th: theta, the current parameter estimates
  
  beta   = th[-length(th)]            # reg coefs to be estimated
  sigma  = th[length(th)]             # sigma to be estimated
  sigma2 = sigma^2
  mu = X %*% beta
  
  # priors are b0 ~ N(0, sd=10), sigma2 ~ invGamma(.001, .001)
  priorbvarinv = diag(1/100, 4) 
  prioralpha   = priorbeta = .001
  
  if (is.nan(sigma) | sigma<=0) {     # scale parameter must be positive, so post
    return(-Inf)                      # density is zero if it jumps below zero
  }
  # log posterior in this conjugate setting. conceptually it's (log) prior +
  # (log) likelihood. (See commented 'else' for alternative)
  else {
    -.5*nrow(X)*log(sigma2) - (.5*(1/sigma2) * (crossprod(y-mu))) +
      -.5*ncol(X)*log(sigma2) - (.5*(1/sigma2) * (t(beta) %*% priorbvarinv %*% beta)) +
      -(prioralpha + 1)*log(sigma2) + log(sigma2) - priorbeta/sigma2
  }
  # else {
  #   ll = mvtnorm::dmvnorm(y, mean=mu, sigma=diag(sigma2, length(y)), log=T)
  #   priorb = mvtnorm::dmvnorm(beta, mean=rep(0, length(beta)), sigma=diag(100, length(beta)), log=T)
  #   priors2 = dgamma(1/sigma2, prioralpha, priorbeta, log=T)
  #   logposterior = ll + priorb + priors2
  #   logposterior
  # }
}
```


The following is the numerical gradient function as given in BDA3 p. 602. It has
the same arguments as the log posterior function.

```{r hmc-func-gradient}
gradient_theta <- function(X, y, th) {
  d = length(th)
  e = .0001
  diffs = numeric(d)
  
  for (k in 1:d) {
    th_hi = th
    th_lo = th
    th_hi[k] = th[k] + e
    th_lo[k] = th[k] - e
    diffs[k] = (log_posterior(X, y, th_hi) - log_posterior(X, y, th_lo)) / (2 * e)
  }
  
  diffs
}
```


The following is a function for a single HMC iteration. ϵ and L are drawn randomly at each
iteration to explore other areas of the posterior (starting with `epsilon0` and
`L0`);  The mass matrix `M`, expressed as a vector, is a bit of a magic number in
this setting. It regards the mass of a particle whose position is represented by
$\theta$, and momentum by $\phi$. See the [sampling section of the Stan
manual](https://mc-stan.org/docs/2_18/reference-manual/hamiltonian-monte-carlo.html)
for more detail.

```{r hmc-func-iter}
hmc_iteration <- function(X, y, th, epsilon, L, M) {
  # Args
  # epsilon: the stepsize
  # L: the number of leapfrog steps
  # M: a diagonal mass matrix

  # initialization
  M_inv = 1/M
  d   = length(th)
  phi = rnorm(d, 0, sqrt(M))
  th_old = th
  
  log_p_old = log_posterior(X, y, th) - .5*sum(M_inv * phi^2)
  
  phi = phi + .5 * epsilon * gradient_theta(X, y, th)
  
  for (l in 1:L) {
    th  = th + epsilon*M_inv*phi
    phi = phi + ifelse(l == L, .5, 1) * epsilon * gradient_theta(X, y, th)
  }
  
  # here we get into standard MCMC stuff, jump or not based on a draw from a
  # proposal distribution
  phi = -phi
  log_p_star = log_posterior(X, y, th) - .5*sum(M_inv * phi^2)    
  r = exp(log_p_star - log_p_old)
  
  if (is.nan(r)) r = 0
  
  p_jump = min(r, 1)
  
  if (runif(1) < p_jump) {
    th_new = th
  }
  else {
    th_new = th_old
  }
  
  # returns estimates and acceptance rate
  list(th = th_new, p_jump = p_jump)  
}
```



Main HMC function.



```{r hmc-func-run}
hmc_run <- function(starts, iter, warmup, epsilon_0, L_0, M, X, y) {
  # # Args: 
  # starts:  starting values
  # iter: total number of simulations for each chain (note chain is based on the dimension of starts)
  # warmup: determines which of the initial iterations will be ignored for inference purposes
  # epsilon0: the baseline stepsize
  # L0: the baseline number of leapfrog steps
  # M: is the mass vector
  chains = nrow(starts)
  d = ncol(starts)
  sims = array(NA, 
               c(iter, chains, d), 
               dimnames = list(NULL, NULL, colnames(starts)))
  p_jump = matrix(NA, iter, chains)
  
  for (j in 1:chains) {
    th = starts[j,]
    
    for (t in 1:iter) {
      epsilon = runif(1, 0, 2*epsilon_0)
      L    = ceiling(2*L_0*runif(1))
      
      temp = hmc_iteration(X, y, th, epsilon, L, M)
      
      p_jump[t,j] = temp$p_jump
      sims[t,j,]  = temp$th
      
      th = temp$th
    }
  }
  
  # acceptance rate
  acc = round(colMeans(p_jump[(warmup + 1):iter,]), 3)  
  
  message('Avg acceptance probability for each chain: ', 
          paste0(acc[1],', ',acc[2]), '\n') 
  
  list(sims = sims, p_jump = p_jump)
}
```

## Estimation

With the primary functions in place, we set the starting values and choose other settings for for the HMC process. The coefficient starting values are based on random draws from a uniform distribution, while $\sigma$ is set to a value of one in each case.  As in the other examples we'll have 5000 total draws with warm-up set to 2500.  I don't have any thinning option here, but that could be added or simply done as part of the <span class="pack">coda</span> package preparation.

```{r hmc-est-initial1}
# Starting values and mcmc settings
parnames = c(paste0('beta[', 1:4, ']'), 'sigma')
d = length(parnames)

chains = 2

theta_start = t(replicate(chains, c(runif(d-1, -1, 1), 1)))
colnames(theta_start) = parnames

nsim = 1000
wu   = 500
```

We can fiddle with these sampling parameters to get a desirable acceptance rate of around .80. The following work well with the data we have.

```{r hmc-est-initial2}
stepsize = .08
nLeap = 10
vars  = rep(1, 5)
mass_vector = 1 / vars
```




We are now ready to run the model.  On my machine and with the above settings, it took a couple seconds. Once complete we can use the <span class="pack">coda</span> package if desired as we have done before.

```{r hmc-est}
# Run the model
fit_hmc = hmc_run(
  starts    = theta_start,
  iter      = nsim,
  warmup    = wu,
  epsilon_0 = stepsize,
  L_0 = nLeap,
  M   = mass_vector,
  X   = X,
  y   = y
)
# str(fit_hmc, 1)
```

Using <span class="pack" style = "">coda</span>, we can get some nice summary information. Results not shown.

```{r hmc-est-coda}
library(coda)

theta = as.mcmc.list(list(as.mcmc(fit_hmc$sims[(wu+1):nsim, 1,]), 
                          as.mcmc(fit_hmc$sims[(wu+1):nsim, 2,])))

# summary(theta)
fit_summary =  summary(theta)$statistics[,'Mean']

beta_est  = fit_summary[1:4]
sigma_est = fit_summary[5]

# log_posterior(X, y, fit_summary)
```


Instead we can use <span class="pack" style = "">rstan's</span> <span class="func" style = "">monitor</span> function on `fit_hmc$sims` to produce typical Stan output.

```{r hmc-est-show, echo=FALSE}
init = rstan::monitor(
  fit_hmc$sims,
  warmup = wu,
  digits_summary = 5,
  probs  = c(.025, .975),
  se     = FALSE,
  print  = F
) %>%
  as_tibble(rownames = 'parameter') 

init %>% 
  select(parameter, mean, sd:Rhat, Bulk_ESS, Tail_ESS) %>% 
  kable_df(caption = glue::glue('log posterior = ', {
  round(log_posterior(X, y, fit_summary), 3)
}))
```

## Comparison

Our estimates look pretty good, and inspection of the diagnostics would show
good mixing and convergence as well. At this point we can compare it to the Stan
output.  For the following, I  use the same inverse gamma prior and tweaked the
control options for a little bit more similarity, but that's not necessary.


```{stan hmc-compare, output.var='stan_hmc'}
data {                            // Data block
  int<lower = 1> N;               // Sample size
  int<lower = 1> K;               // Dimension of model matrix
  matrix [N, K]  X;               // Model Matrix
  vector[N] y;                    // Target variable
}

parameters {                      // Parameters block; declarations only
  vector[K] beta;                 // Coefficient vector
  real<lower = 0> sigma;          // Error scale
}

model {                           // Model block
  vector[N] mu;

  mu = X * beta;                  // Creation of linear predictor
  
  
  // priors
  beta  ~ normal(0, 10);
  sigma ~ inv_gamma(.001, .001);  // changed to gamma a la code above
  
  // likelihood
  y ~ normal(mu, sigma);
}
```


```{r hmc-stan-est, results='hide', echo=FALSE}
dat = list(N = N,
           K = ncol(X),
           y = y,
           X = X)

library(rstan)

# standard stan
# fit = stan(
#   model_code = stan_hmc_code_as_string,
#   data = dat,
#   iter = nsim,
#   warmup = wu,
#   thin = 1,
#   chains = chains,
#   verbose = F
# )

# perhaps closer to above settings?
fit_stan = sampling(
  stan_hmc,
  data     = dat,
  iter     = nsim,
  warmup   = wu,
  thin     = 1,
  chains   = chains,
  verbose  = FALSE,
  control  = list(
    adapt_engaged = FALSE,
    stepsize      = stepsize,
    adapt_t0      = nLeap
  )
)
```

Here are the results.

```{r hmc-compare-stan, echo=FALSE}
broom.mixed::tidy(fit_stan, conf.int = TRUE) %>%
  kable_df(
    caption = glue::glue('log posterior = ', round(get_posterior_mean(fit_stan, par='lp__')[,3],3))
  )
```


And finally, the standard least squares fit (Residual standard error = sigma).

```{r hmc-compare-lm, echo=FALSE}
pander::pander(summary(fit_lm))
```

## Source

Original demo here:

https://m-clark.github.io/bayesian-basics/appendix.html#hamiltonian-monte-carlo-example

