# Hurdle Model


Hurdle models are applied to situations in which target data has relatively many of one value, usually zero, to go along with the other observed values.  They are two-part models, a logistic model for whether an observation is zero or not, and a count model for the other part. The key distinction from the usual 'zero-inflated' count models, is that the count distribution does not contribute to the excess zeros.  While the typical application is count data, the approach can be applied to any distribution in theory. 

## Poisson

### Data Setup

Here we import a simple data set. The example comes from the Stata help file for zinb command. One can compare results with `hnblogit` command in Stata.

```{r hurdle-setup}
library(tidyverse)

fish = haven::read_dta("http://www.stata-press.com/data/r11/fish.dta")
```

### Function

The likelihood function is of two parts, one a logistic model, the other, a poisson count model.

```{r pois-hurdle-ll}
hurdle_poisson_ll <- function(y, X, par) {
  # Extract parameters
  logitpars = par[grep('logit', names(par))]
  poispars  = par[grep('pois', names(par))]
  
  # Logit model part
  Xlogit = X
  ylogit = ifelse(y == 0, 0, 1)
  
  LPlogit = Xlogit %*% logitpars
  mulogit = plogis(LPlogit)
  
  # Calculate the likelihood
  logliklogit = -sum( ylogit*log(mulogit) + (1 - ylogit)*log(1 - mulogit) )  
  
  # Poisson part
  Xpois = X[y > 0, ]
  ypois = y[y > 0]
  
  mupois = exp(Xpois %*% poispars)
  
  # Calculate the likelihood
  loglik0    = -mupois
  loglikpois = -sum(dpois(ypois, lambda = mupois, log = TRUE)) + sum(log(1 - exp(loglik0)))
  
  # combine likelihoods
  loglik = loglikpois + logliklogit
  loglik
}
```




Get some starting values from <span class="func" style = "">glm</span> For these functions, and create a named vector for them.

```{r pois-hurdle-starts}
init_mod = glm(
  count ~ persons + livebait,
  data   = fish,
  family = poisson,
  x = TRUE,
  y = TRUE
)

starts = c(logit = coef(init_mod), pois = coef(init_mod))  
```


### Estimation

Use <span class="func" style = "">optim</span>. to estimate parameters. I fiddle with some options to reproduce the  hurdle function as much as possible.
 
```{r pois-hurdle-est}
fit = optim(
  par = starts,
  fn  = hurdle_poisson_ll,
  X   = init_mod$x,
  y   = init_mod$y,
  control = list(maxit = 5000, reltol = 1e-12),
  hessian = TRUE
)
# fit
```



Extract the elements from the output to create a summary table.

```{r pois-hurdle-ext}
B  = fit$par
se = sqrt(diag(solve(fit$hessian)))
Z  = B/se
p  = ifelse(Z >= 0, pnorm(Z, lower = FALSE)*2, pnorm(Z)*2)
summary_table = round(data.frame(B, se, Z, p), 3)

list(summary = summary_table, ll = fit$value)
```

### Comparison

Compare to <span class="func" style = "">hurdle</span> from <span class="pack" style = "">pscl</span> package.

```{r pois-hurdle-pscl}
library(pscl)

fit_pscl = hurdle(
  count ~ persons + livebait,
  data = fish,
  zero.dist = "binomial",
  dist = "poisson"
)
```


```{r pois-hurdle-pscl-show, echo=FALSE}
init1 = purrr::map(summary(fit_pscl)$coefficients, function(x) {
  data.frame(x)
  colnames(x) = colnames(summary_table)
  x
}) %>%
  do.call(rbind, .) %>%
  as.data.frame() %>%
  rownames_to_column('coef') 

init2 = summary_table %>% rownames_to_column('coef')

bind_rows(list(pscl = init1, hurdle_poisson_ll = init2), .id = '') %>% 
  kable_df()
```




## Negative Binomial

### Function

The likelihood function.

```{r nb-hurdle-ll}
hurdle_nb_ll <- function(y, X, par) {
  # Extract parameters
  logitpars  = par[grep('logit', names(par))]
  NegBinpars = par[grep('NegBin', names(par))]
  
  theta = exp(par[grep('theta', names(par))])
  
  # Logit model part
  Xlogit = X
  ylogit = ifelse(y == 0, 0, 1)
  
  LPlogit = Xlogit%*%logitpars
  mulogit =  plogis(LPlogit)
  
  # Calculate the likelihood
  logliklogit = -sum( ylogit*log(mulogit) + (1 - ylogit)*log(1 - mulogit) )
  
  #NB part
  XNB = X[y > 0, ]
  yNB = y[y > 0]
  
  muNB = exp(XNB %*% NegBinpars)
  
  # Calculate the likelihood
  loglik0  = dnbinom(0,   mu = muNB, size = theta, log = TRUE)
  loglik1  = dnbinom(yNB, mu = muNB, size = theta, log = TRUE)
  loglikNB = -( sum(loglik1) - sum(log(1 - exp(loglik0))) )
  
  # combine likelihoods
  loglik = loglikNB + logliklogit
  loglik
}
```

### Estimation

```{r nb-hurdle-est}
starts =  c(
  logit  = coef(init_mod),
  NegBin = coef(init_mod),
  theta  = 1
)

optNB1 = optim(
  par = starts,
  fn  = hurdle_nb_ll,
  X   = init_mod$x,
  y   = init_mod$y,
  control = list(maxit = 5000, reltol = 1e-12),
  method  = "BFGS",
  hessian = TRUE
)
# optNB1 

B  = optNB1$par
se = sqrt(diag(solve(optNB1$hessian)))
Z  = B/se
p  = ifelse(Z >= 0, pnorm(Z, lower = FALSE)*2, pnorm(Z)*2)

summary_table = round(data.frame(B, se, Z, p), 3)

list(summary = summary_table, ll = optNB1$value)
```



### Comparison


```{r nb-hurdle-compare}
fit_pscl = hurdle(
  count ~ persons + livebait,
  data = fish,
  zero.dist = "binomial",
  dist = "negbin"
)

# summary(fit_pscl)$coefficients
# summary_table
```

```{r nb-hurdle-pscl-show, echo=FALSE}
init1 = purrr::map(summary(fit_pscl)$coefficients, function(x) {
  data.frame(x)
  colnames(x) = colnames(summary_table)
  x
}) %>%
  do.call(rbind, .) %>%
  as.data.frame() %>%
  rownames_to_column('coef') 

init2 = summary_table %>% rownames_to_column('coef')

bind_rows(list(pscl = init1, hurdle_nb_ll = init2), .id = '') %>% 
  kable_df()
```



## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/hurdle.R
