# Logistic Regression


The following demo regards a standard logistic regression model via maximum
likelihood or exponential loss. This can serve as an entry point for those
starting out to the wider world of computational statistics as maximum
likelihood is the fundamental approach used in most applied statistics, but
which is also a key aspect of the Bayesian approach.  Exponential loss is not
confined to the standard GLM setting, but is widely used in more
predictive/'algorithmic' approaches e.g. in machine learning and elsewhere.

This follows the [linear regression model][Linear Regression] approach.



## Data Setup 

Predictors and target.

```{r data-setup-logreg}
library(tidyverse)

set.seed(1235)  # ensures replication

N = 10000 # sample size
k = 2     # number of desired predictors
X = matrix(rnorm(N * k), ncol = k)

# the linear predictor
lp = -.5 + .2 * X[, 1] + .1 * X[, 2] # increasing N will get estimated values closer to these

y = rbinom(N, size = 1, prob = plogis(lp))

dfXy = data.frame(X, y)
```

## Functions 
 
A maximum likelihood approach.


```{r logreg_ML}
logreg_ML <- function(par, X, y) {
  # arguments- 
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  
  # setup
  beta = par                                # coefficients
  N = nrow(X)
  
  # linear predictor
  LP = X %*% beta                           # linear predictor
  mu = plogis(LP)                           # logit link
  
  # calculate likelihood
  L = dbinom(y, size = 1, prob = mu, log = TRUE)         # log likelihood
  #   L =  y*log(mu) + (1 - y)*log(1-mu)    # alternate log likelihood form
  
  -sum(L)                                   # optim by default is minimization, and we want to maximize the likelihood 
  # (see also fnscale in optim.control)
}
```



Another approach via exponential loss function.

```{r logreg_exp}
logreg_exp <- function(par, X, y) {
  # arguments- 
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  
  # setup
  beta = par                                   # coefficients
  
  # linear predictor
  LP = X %*% beta                              # linear predictor

  # calculate exponential loss function (convert y to -1:1 from 0:1)
  L = sum(exp(-ifelse(y, 1, -1) * .5 * LP))
}
```




## Estimation

Setup for use with <span class="func" style = "">optim</span>.

```{r logreg-est}
X = cbind(1, X)

# initial values

init = rep(0, ncol(X))
names(init) = c('intercept', 'b1', 'b2')

optlmML = optim(
  par = init,
  fn  = logreg_ML,
  X   = X,
  y   = y,
  control = list(reltol = 1e-8)
)

optglmClass = optim(
  par = init,
  fn  = logreg_exp,
  X   = X,
  y   = y, 
  control = list(reltol = 1e-15)
)

pars_ML  = optlmML$par
pars_exp = optglmClass$par
```





## Comparison
 
Compare to `glm`.

```{r logreg-compare}
modglm = glm(y ~ ., dfXy, family = binomial)

rbind(
  pars_ML,
  pars_exp,
  pars_GLM = coef(modglm)
)

```



## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/standard_logistic.R
