# Two-factor Mixed Model



An approach for two factor random effects model via maximum likelihood in R Matlab and Julia. It's based on Statistical Modeling and Computation (2014) Chapter 10, example 10.10. The data regards the breeding value of a set of five sires in raising pigs. Each sire is mated to a random group of dams, with the response being the average daily weight gain in pounds of two piglets in each litter.  See the [previous chapter][One-factor Mixed Model] for a one factor model, and two_factor_RE.m and two_factor_RE.jl for the Matlab and Julia versions of this example on the GitHub site. Note that the text has a typo on the `sigma2` variance estimate (value should be .0023 not .023).


## Data Setup

```{r two-factor-re-setup}
library(tidyverse)

y = c(1.39,1.29,1.12,1.16,1.52,1.62,1.88,1.87,1.24,1.18,
      .95,.96,.82,.92,1.18,1.20,1.47,1.41,1.57,1.65)

# for use in lme4, but also a more conceptual representation of the data
d = expand.grid(sire = rep(1:5, 2), dam = 1:2)
d = data.frame(d[order(d$sire), ], y)
```


## Function

The function takes the log variances `eta`* as input to keep positive.

```{r two-factor-re}
two_factor_re = function(mu, eta_alpha, eta_gamma, eta) {
  # Args 
  # mu: intercept 
  # eta_alpha: random effect one 
  # eta_gamma: random effect two
  # eta: residual variance of y
  
  sigma2_alpha = exp(eta_alpha)
  sigma2_gamma = exp(eta_gamma)
  sigma2 = exp(eta)
  n = length(y)
  
  # covariance matrix of observations
  Sigma = sigma2 * diag(n) + sigma2_alpha * tcrossprod(Xalpha) + 
    sigma2_gamma * tcrossprod(Xgamma)
  
  
  # log likelihood
  ll = -n / 2 * log(2 * pi) - sum(log(diag(chol(Sigma)))) -
    .5 * t(y - mu) %*% chol2inv(chol(Sigma)) %*% (y - mu)
  
  return(-ll)
}
```




## Estimation

Starting values and test.

```{r two-factor-re-starts}
starts = list(
  mu = mean(y),
  eta_alpha = var(tapply(y, d$sire, mean)),
  eta_gamma = var(y) / 3,
  eta = var(y) / 3
)

Xalpha = diag(5) %x% rep(1,4)

Xgamma = diag(10) %x% rep(1,2)
```



Estimation.

```{r two-factor-re-est}
two_factor_re(starts[[1]], starts[[2]], starts[[3]], starts[[4]])
```




## Comparison


```{r two-factor-re-compare}
library(bbmle)

mlout = mle2(two_factor_re, start=starts,  method='BFGS')  

### lme4 comparison
library(lme4)

lme = lmer(y ~ (1 | sire) + (1 | dam:sire), d, REML = F)

summary(mlout)
exp(coef(mlout)[-1])

summary(lme)
```


## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/Mixed%20Models/two_factor_RE.R




