# Nelder-Mead

This is based on the pure [Python implementation by François
Chollet](https://github.com/fchollet/nelder-mead), also found in the
[supplemental section](#python-nelder). This is mostly just an academic exercise
on my part.  I'm not sure how much one would use the basic NM for many
situations. In my experience BFGS and other approaches would be faster, more
accurate, and less sensitive to starting values for the types of problems I've
played around with. Others who actually spend their time researching such things
seem to agree.

There were two issues  regarding the original code on
[GitHub](https://github.com/fchollet/nelder-mead/issues/2), and I've implemented
the suggested corrections with notes. The initial R function code is not very
R-like, as the goal was to keep more similar to the original Python for
comparison, which used a list approach. I also provide a more R-like/cleaner
version that uses matrices instead of lists, but which still sticks the same
approach for the most part.
For both functions, comparisons are made using the <span class="pack" style =
"">optimx</span> package, but feel free to use base R's <span class="func" style
= "">optim</span> instead.




## Functions

### First Version

 - `f`  function to optimize, must return a scalar score and operate over
  an array of the same dimensions as x_start
 - `x_start` initial position
 - `step` look-around radius in initial step
 - `no_improve_thr` See no_improv_break
 - `no_improv_break` break after no_improv_break iterations with an
  improvement lower than no_improv_thr
 - `max_iter` always break after this number of iterations. Set it to 0 to
  loop indefinitely.
 - `alpha` parameters of the algorithm (see Wikipedia page for reference)
 - `gamma` parameters of the algorithm (see Wikipedia page for reference)
 - `rho` parameters of the algorithm (see Wikipedia page for reference)
 - `sigma` parameters of the algorithm (see Wikipedia page for reference)
 - `verbose` Print iterations?

This function returns the best parameter array and best score.

```{r nelder-mead}
nelder_mead <- function(
  f, 
  x_start,
  step = 0.1,
  no_improve_thr  = 1e-12,
  no_improv_break = 10,
  max_iter = 0,
  alpha    = 1,
  gamma    = 2,
  rho      = 0.5,
  sigma    = 0.5,
  verbose  = FALSE
  ) {
  # init
  dim = length(x_start)
  prev_best = f(x_start)
  no_improv = 0
  res = list(list(x_start = x_start, prev_best = prev_best))
  
  
  for (i in 1:dim) {
    x = x_start
    x[i]  = x[i] + step
    score = f(x)
    res = append(res, list(list(x_start = x, prev_best = score)))
  }
  
  # simplex iter
  iters = 0
  
  while (TRUE) {
    # order
    idx  = sapply(res, `[[`, 2)
    res  = res[order(idx)]   # ascending order
    best = res[[1]][[2]]
    
    # break after max_iter
    if (max_iter > 0 & iters >= max_iter) return(res[[1]])
    iters = iters + 1
    
    # break after no_improv_break iterations with no improvement
    if (verbose) message(paste('...best so far:', best))
    
    if (best < (prev_best - no_improve_thr)) {
      no_improv = 0
      prev_best = best
    } else {
      no_improv = no_improv + 1
    }
    
    if (no_improv >= no_improv_break) return(res[[1]])
    
    # centroid
    x0 = rep(0, dim)
    for (tup in 1:(length(res)-1)) {
      for (i in 1:dim) {
        x0[i] = x0[i] + res[[tup]][[1]][i] / (length(res)-1)
      }
    }
    
   # reflection
   xr = x0 + alpha * (x0 - res[[length(res)]][[1]])
   rscore = f(xr)
   if (res[[1]][[2]] <= rscore & 
       rscore < res[[length(res)-1]][[2]]) {
     res[[length(res)]] = list(xr, rscore)
     next
   }
     
   # expansion
   if (rscore < res[[1]][[2]]) {
     # xe = x0 + gamma*(x0 - res[[length(res)]][[1]])   # issue with this
     xe = x0 + gamma * (xr - x0)   
     escore = f(xe)
     if (escore < rscore) {
       res[[length(res)]] = list(xe, escore)
       next
     } else {
       res[[length(res)]] = list(xr, rscore)
       next
     }
   }
   
   # contraction
   # xc = x0 + rho*(x0 - res[[length(res)]][[1]])  # issue with wiki consistency for rho values (and optim)
   xc = x0 + rho * (res[[length(res)]][[1]] - x0)
   cscore = f(xc)
   if (cscore < res[[length(res)]][[2]]) {
     res[[length(res)]] = list(xc, cscore)
     next
   }
   
   # reduction
   x1   = res[[1]][[1]]
   nres = list()
   for (tup in res) {
     redx  = x1 + sigma * (tup[[1]] - x1)
     score = f(redx)
     nres  = append(nres, list(list(redx, score)))
   }
   
   res = nres
  }
  
  res
}
```






### Example

The function to minimize.

```{r opt-func-1}
f = function(x) {
  sin(x[1]) * cos(x[2]) * (1 / (abs(x[3]) + 1))
}
```

Estimate.

```{r nelder-mead-est-1}
nelder_mead(
  f, 
  c(0, 0, 0), 
  max_iter = 1000, 
  no_improve_thr = 1e-12
)
```





Compare to <span class="pack" style = "">optimx</span>.  You may see warnings.

```{r nelder-mead-compare-optimx-1}
optimx::optimx(
  par = c(0, 0, 0),
  fn = f,
  method = "Nelder-Mead",
  control = list(
    alpha  = 1,
    gamma  = 2,
    beta   = 0.5,
    maxit  = 1000,
    reltol = 1e-12
  )
)
```




### A Regression Model

I find a regression model to be more applicable/intuitive for my needs, so
provide an example for that case.


#### Data Setup

```{r nelder-mead-setup-1}
library(tidyverse)

set.seed(8675309)

N = 500
n_preds = 5  # number of predictors
X = cbind(1, matrix(rnorm(N * n_preds), ncol = n_preds))
beta = runif(ncol(X), -1, 1)
y = X %*% beta + rnorm(nrow(X))
```



Least squares loss function to minimize.

```{r nelder-mead-compare-ls-1}
f = function(b) {
  crossprod(y - X %*% b)[,1]  # if using optimx need scalar
}
```




```{r nelder-mead-est-lm-1}
fit_nm = nelder_mead(
  f, 
  runif(ncol(X)), 
  max_iter = 2000,
  no_improve_thr = 1e-12,
  verbose = FALSE
)
```



#### Comparison

Compare to <span class="pack" style = "">optimx</span>.

```{r nelder-mead-compare-optimx}
fit_nm_optimx = optimx::optimx(
  runif(ncol(X)),
  fn = f,  # model function
  method  = 'Nelder-Mead',
  control = list(
    alpha = 1,
    gamma = 2,
    beta  = 0.5,
    #rho
    maxit  = 2000,
    reltol = 1e-12
  )
)
```

We'll add base R <span class="func" style = "">lm</span> estimates.

```{r nelder-mead-compare-1, echo=FALSE}
fit_lm = lm.fit(X, y)

rbind(
  nm_func   = unlist(fit_nm),
  nm_optimx = fit_nm_optimx[1:7],
  lm = c(fit_lm$coef, sum(resid(fit_lm)^2))
) %>% 
  kable_df()
```




## Second Version

This is a more natural R approach in my opinion.

```{r nelder-mead-2}
nelder_mead2 = function(
  f,
  x_start,
  step = 0.1,
  no_improve_thr  = 1e-12,
  no_improv_break = 10,
  max_iter = 0,
  alpha    = 1,
  gamma    = 2,
  rho      = 0.5,
  sigma    = 0.5,
  verbose  = FALSE
) {
  
  # init
  npar = length(x_start)
  nc = npar + 1
  prev_best = f(x_start)
  no_improv = 0
  res = matrix(c(x_start, prev_best), ncol = nc)
  colnames(res) = c(paste('par', 1:npar, sep = '_'), 'score')
  
  for (i in 1:npar) {
    x     = x_start
    x[i]  = x[i] + step
    score = f(x)
    res   = rbind(res, c(x, score))
  }
  
  # simplex iter
  iters = 0
  
  while (TRUE) {
    # order
    res  = res[order(res[, nc]), ]   # ascending order
    best = res[1, nc]
    
    # break after max_iter
    if (max_iter & iters >= max_iter) return(res[1, ])
    iters = iters + 1
    
    # break after no_improv_break iterations with no improvement
    if (verbose) message(paste('...best so far:', best))
    
    if (best < (prev_best - no_improve_thr)) {
      no_improv = 0
      prev_best = best
    } else {
      no_improv = no_improv + 1
    }
    
    if (no_improv >= no_improv_break)
      return(res[1, ])
    
    nr = nrow(res)
    
    # centroid: more efficient than previous double loop
    x0 = colMeans(res[(1:npar), -nc])
    
    # reflection
    xr = x0 + alpha * (x0 - res[nr, -nc])
    
    rscore = f(xr)
    
    if (res[1, 'score'] <= rscore & rscore < res[npar, 'score']) {
      res[nr,] = c(xr, rscore)
      next
    }
    
    # expansion
    if (rscore < res[1, 'score']) {
      xe = x0 + gamma * (xr - x0)
      escore = f(xe)
      if (escore < rscore) {
        res[nr, ] = c(xe, escore)
        next
      } else {
        res[nr, ] = c(xr, rscore)
        next
      }
    }
    
    # contraction
    xc = x0 + rho * (res[nr, -nc] - x0)
    
    cscore = f(xc)
    
    if (cscore < res[nr, 'score']) {
      res[nr,] = c(xc, cscore)
      next
    }
    
    # reduction
    x1 = res[1, -nc]
    
    nres = res
    
    for (i in 1:nr) {
      redx  = x1 + sigma * (res[i, -nc] - x1)
      score = f(redx)
      nres[i, ] = c(redx, score)
    }
    
    res = nres
  }
}
```




### Example

The function to minimize.

```{r opt-func-2}
f = function(x) {
  sin(x[1]) * cos(x[2]) * (1 / (abs(x[3]) + 1))
}
```

```{r nelder-mead-compare-optimx-2}
nelder_mead2(
  f, 
  c(0, 0, 0), 
  max_iter = 1000, 
  no_improve_thr = 1e-12
)

optimx::optimx(
  par = c(0, 0, 0), 
  fn = f, 
  method   = "Nelder-Mead",
  control  = list(
    alpha  = 1,
    gamma  = 2,
    beta   = 0.5,
    maxit  = 1000,
    reltol = 1e-12
  )
)
```






### A Regression Model

```{r nelder-mead-setup-2}
set.seed(8675309)

N = 500
n_preds = 5
X = cbind(1, matrix(rnorm(N * n_preds), ncol = n_preds))
beta = runif(ncol(X), -1, 1)
y = X %*% beta + rnorm(nrow(X))
```


Least squares loss function to minimize.


```{r nelder-mead-compare-ls-2}
f = function(b) {
  crossprod(y - X %*% b)[,1]  # if using optimx need scalar
}
```




```{r nelder-mead-est-2}
fit_nm = nelder_mead2(
  f, 
  runif(ncol(X)),
  max_iter = 2000,
  no_improve_thr = 1e-12
)
```

#### Comparison

Compare to <span class="pack" style = "">optimx</span> and <span class="func" style = "">lm</span> as before.

```{r nelder-mead-compare-2}
fit_nm_optimx = optimx::optimx(
  runif(ncol(X)),
  fn = f,
  method   = 'Nelder-Mead',
  control  = list(
    alpha  = 1,
    gamma  = 2,
    beta   = 0.5,
    maxit  = 2000,
    reltol = 1e-12
  )
)[1:(n_preds + 1)]

fit_lm = lm.fit(X, y)$coef
```

```{r nelder-mead-compare-2-show, echo=FALSE}
rbind(
  nm_func   = unlist(fit_nm),
  nm_optimx = fit_nm_optimx,
  lm    = fit_lm,
  truth = beta
) %>% 
  t() %>% 
  kable_df()
```




## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/nelder_mead.R
