# Ordinal

The following demonstrates a standard cumulative link ordinal regression model
via maximum likelihood. Default is with probit link function. Alternatively you
can compare it with a logit link, which will result in values roughly
1.7*parameters estimates from the probit.


## Data

This data generation is from the probit perspective, where the underlying continuous latent variable is normally distributed.

```{r ordinal-setup}
library(tidyverse)

set.seed(808)

N = 1000                                       # Sample size
x = cbind(x1 = rnorm(N), x2 = rnorm(N))        # predictor variables
beta   = c(1,-1)                               # coefficients
y_star = rnorm(N, mean = x %*% beta)           # the underlying latent variable
y_1 = y_star > -1.5                            # -1.50 first cutpoint
y_2 = y_star > .75                             #  0.75 second cutpoint
y_3 = y_star > 1.75                            #  1.75 third cutpoint
y   = y_1 + y_2 + y_3 + 1                      # target

table(y)

d = data.frame(x, y = factor(y))
```


## Function

```{r ordinal_ll}
ordinal_ll <- function(par, X, y, probit = TRUE) {
  K = length(unique(y))                        # number of classes K
  ncuts = K-1                                  # number of cutpoints/thresholds
  cuts  = par[(1:ncuts)]                       # cutpoints
  beta  = par[-(1:ncuts)]                      # regression coefficients
  lp    = X %*% beta                           # linear predictor
  ll    = rep(0, length(y))                    # log likelihood
  pfun  = ifelse(probit, pnorm, plogis)        # which link to use
  
  for(k in 1:K){
    if (k==1) {
      ll[y==k] = pfun((cuts[k] - lp[y==k]), log = TRUE)
    }
    else if (k < K) {
      ll[y==k] = log(pfun(cuts[k] - lp[y==k]) - pfun(cuts[k-1] - lp[y==k]))
    }
    else {
      ll[y==k] = log(1 - pfun(cuts[k-1] - lp[y==k])) 
    }
  }
  
  -sum(ll)
}
```


## Estimation

```{r ordinal-init}
init = c(-1, 1, 2, 0, 0)         # initial values
```

```{r ordinal-est}
fit_probit = optim(
  init,
  ordinal_ll,
  y = y,
  X = x,
  probit  = TRUE,
  control = list(reltol = 1e-10)
)

fit_logit  = optim(
  init,
  ordinal_ll,
  y = y,
  X = x,
  probit  = FALSE,
  control = list(reltol = 1e-10)
)
```




## Comparison

We can compare our results with the <span class="pack" style = "">ordinal</span> package.

```{r ordinal-compare}
library(ordinal)
fit_ordpack_probit = clm(y ~ x1 + x2, data = d, link = 'probit')
fit_ordpack_logit  = clm(y ~ x1 + x2, data = d, link = 'logit')
```


```{r ordinal-compare-show, echo=FALSE}
resprobit = data.frame(method = c('ordinal_ll', 'ordpack'),
                       rbind(coef(fit_probit), coef(fit_ordpack_probit)))
colnames(resprobit) = c('method', paste0('cut_', 1:3), 'beta1', 'beta2')



reslogit = data.frame(method = c('ordinal_ll', 'ordpack'),
                      rbind(coef(fit_logit), coef(fit_ordpack_logit)))
colnames(reslogit) = c('method', paste0('cut_', 1:3), 'beta1', 'beta2')

bind_rows(list(probit = resprobit, logit = reslogit), .id = '') %>% 
  kable_df()
```




## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/ordinal_regression.R
