# Penalized Maximum Likelihood



This demonstration regards a standard regression model via penalized likelihood.
See the [Maximum Likelihood chapter][Maximum Likelihood] for a starting point. Here
the penalty is specified (via `lambda` argument), but one would typically
estimate the model via cross-validation or some other fashion. Two penalties are
possible with the function.  One using the (squared) L2 norm (aka ridge
regression, Tikhonov regularization), another using the L1 norm (aka lasso)
which has the possibility of penalizing coefficients to zero, and thus can serve
as a model selection procedure. I have more technical approaches to the lasso
and ridge in the [lasso][L1 (lasso) regularization] and [ridge][L2 (ridge)
regularization] chapters.

Note that both L2 and L1 approaches can be seen as maximum a posteriori (MAP)
estimates for a Bayesian regression with a specific prior on the
coefficients. The L2 approach is akin to a normal prior with zero mean, while
L1 is akin to a zero mean Laplace prior.  See the [Bayesian regression chapter][Bayesian Linear Regression] for an approach in that regard.




##  Data Setup

```{r pml-data-setup}
library(tidyverse)

set.seed(123)  # ensures replication

# predictors and response
N = 100 # sample size
k = 2   # number of desired predictors
X = matrix(rnorm(N * k), ncol = k)  
y = -.5 + .2*X[, 1] + .1*X[, 2] + rnorm(N, sd = .5)  # increasing N will get estimated values closer to these

dfXy = data.frame(X,y)
```


## Functions

A straightforward function to estimate the log likelihood.

```{r penalized_ML}
penalized_ML <- function(
  par,
  X,
  y,
  lambda = .1,
  type = 'L2'
) {
  # arguments- 
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  # lambda: penalty coefficient
  # type: penalty approach
  
  # setup
  beta   = par[-1]                               # coefficients
  sigma2 = par[1]                                # error variance
  sigma  = sqrt(sigma2)
  N = nrow(X)

  LP = X %*% beta                                # linear predictor
  mu = LP                                        # identity link in the glm sense
  
  # calculate likelihood
  L = dnorm(y, mean = mu, sd = sigma, log = T)   # log likelihood
  
  PL = switch(type,
              'L2' = -sum(L) + lambda * crossprod(beta[-1]),   # the intercept is not penalized
              'L1' = -sum(L) + lambda * sum(abs(beta[-1]))
              )
}
```



This next function is a <span class="pack" style = "">glmnet</span>  style
approach that will put the lambda coefficient on equivalent scale. It uses a
different objective function.  Note that <span class="pack" style =
"">glmnet</span> is actually using `elasticnet`, which mixes both L1 and L2
penalties.

```{r penalized_ML2}
penalized_ML2 <- function(
  par,
  X,
  y,
  lambda = .1,
  type = 'L2'
) {
  
  # arguments- 
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  # lambda: penalty coefficient
  # type: penalty approach
  
  # setup
  beta = par                                   # coefficients
  N = nrow(X)
  
  # linear predictor
  LP = X %*% beta                              # linear predictor
  mu = LP                                      # identity link in the glm sense
  
  obj = switch(type,
               'L2' = .5*crossprod(y - X %*% beta)/N + lambda * crossprod(beta[-1]),
               'L1' = .5*crossprod(y - X %*% beta)/N + lambda * sum(abs(beta[-1]))
  )
}
```



## Estimation

Setup the model matrix for use with <span class="func" style = "">optim</span>.

```{r penalized-ml-mm}
X = cbind(1, X)
```


We'll need to set initial values.  Note we'd normally want to handle the sigma
parameter differently as it's bounded by zero, but we'll ignore for
demonstration.

```{r penalized-ml-est}
init = c(1, rep(0, ncol(X)))
names(init) = c('sigma2', 'intercept', 'b1', 'b2')

optlmpenalized_MLL2 = optim(
  par = init,
  fn  = penalized_ML,
  X   = X,
  y   = y,
  lambda  = 1,
  control = list(reltol = 1e-12)
)

optlmpenalized_MLL1 = optim(
  par = init,
  fn  = penalized_ML,
  X   = X,
  y   = y,
  lambda  = 1,
  type    = 'L1',
  control = list(reltol = 1e-12)
)

parspenalized_MLL2 = optlmpenalized_MLL2$par
parspenalized_MLL1 = optlmpenalized_MLL1$par
```


## Comparison

Compare to `lm` in base R.

```{r penalized-ml-compare}
modlm = lm(y ~ ., dfXy)

round(
  rbind(
    parspenalized_MLL2,
    parspenalized_MLL1,
    modlm = c(summary(modlm)$sigma ^ 2, coef(modlm))
  ), 
  digits = 4
)
```



Compare to <span class="pack" style = "">glmnet</span>. Setting alpha to 0 and 1
is equivalent to L2 and L1 penalties respectively. You also wouldn't want to
specify lambda normally, and rather let it come about as part of the estimation
procedure.  We do so here just for demonstration.


```{r penalized-ml-compare-glmnet}
library(glmnet)

glmnetL2 = glmnet(
  X[, -1],
  y,
  alpha  = 0,
  lambda = .01,
  standardize = FALSE
)

glmnetL1 = glmnet(
  X[, -1],
  y,
  alpha  = 1,
  lambda = .01,
  standardize = FALSE
)

pars_L2 = optim(
  par = init[-1],
  fn  = penalized_ML2,
  X   = X,
  y   = y,
  lambda  = .01,
  control = list(reltol = 1e-12)
)$par

pars_L1 = optim(
  par = init[-1],
  fn  = penalized_ML2,
  X   = X,
  y   = y,
  lambda  = .01,
  type    = 'L1',
  control = list(reltol = 1e-12)
)$par
```


```{r penalized-ml-compare-glmnet-show, echo=FALSE}
rbind(
    glmnet_L2 = t(as.matrix(coef(glmnetL2))),
    pars_L2   = pars_L2,
    glmnet_L1 = t(as.matrix(coef(glmnetL1))),
    pars_L1   = pars_L1
  ) %>% 
  kable_df(digits = 4)
```




## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/penalized_ML.R
