# Probit & Bivariate Probit

Stata users seem to be the primary audience concerned with probit models, but I
thought I'd play around with one even though I've never had reason to use it.
Stata examples come from the UCLA ATS website and the Stata manual, so one can
investigate the Stata result for comparison.

## Standard Probit

The standard probit model is identical to the [logistic model][Logistic
Regression] but using a different link function.

### Function

```{r probit_ll}
probit_ll = function(beta, X, y){
  mu = X %*% beta
  
  # these produce identical results, but the second is the typical depiction
  ll = sum(dbinom(
    y,
    size = 1,
    prob = pnorm(mu),
    log = T
  ))
  
  # ll = sum(y * pnorm(mu, log = T) + (1 - y) * log(1 - pnorm(mu)))
  
  -ll
}
```


### Examples

Example 1 available at https://stats.idre.ucla.edu/stata/dae/probit-regression/

```{r probit-example-1}
library(tidyverse)

admit = haven::read_dta('https://stats.idre.ucla.edu/stat/stata/dae/binary.dta')

head(admit)

X = model.matrix(admit~ gre + gpa + factor(rank), admit)
y = admit$admit

init = rep(0, ncol(X))

fit = optim(
  fn  = probit_ll,
  par = init,
  X = X,
  y = y,
  method = 'BFGS'
)

fit 
```




Example 2 from Stata manual on [standard probit](http://www.stata.com/manuals13/rprobit.pdf) 

> We have data on the make, weight, and mileage rating of 22 foreign and 52 domestic automobiles. We wish to fit a probit model explaining whether a car is foreign based on its weight and mileage."

```{r probit-example-2}
auto = haven::read_dta('http://www.stata-press.com/data/r13/auto.dta')

head(auto)

X = model.matrix(foreign~ weight + mpg, auto)
y = auto$foreign

init = rep(0, ncol(X))

fit = optim(
  fn  = probit_ll,
  par = init,
  X = X,
  y = y
)

fit
```



## Bivariate Probit

For the bivariate model, we are dealing with two binary outcomes and their correlation.


Here is the main function.

```{r bivariate_probit_ll}
bivariate_probit_ll = function(pars, X, y1, y2) {
  rho = pars[1]
  mu1 = X %*% pars[2:(ncol(X) + 1)]
  mu2 = X %*% pars[(ncol(X) + 2):length(pars)]
  q1  = ifelse(y1 == 1, 1,-1)
  q2  = ifelse(y2 == 1, 1,-1)
  
  require(mnormt)
  eta1 = q1 * mu1
  eta2 = q2 * mu2
  
  ll = matrix(NA, nrow = nrow(X))
  
  for (i in 1:length(ll)) {
    corr = q1[i] * q2[i] * rho
    corr = matrix(c(1, corr, corr, 1), 2)
    ll[i] = log(
      pmnorm(
        x = c(eta1[i], eta2[i]),
        mean   = c(0, 0),
        varcov = corr
      )
    )
  }
  
  # the loop is probably clearer, and there is no difference in time, but here's
  # a oneliner ll = mapply(function(e1, e2, q1, q2) log(pmnorm(x = c(e1, e2),
  # varcov = matrix(c(1,q1*q2*rho,q1*q2*rho,1),2))), eta1, eta2, q1, q2)
  
  -sum(ll)
}
```




### Example 

From the Stata manual on bivariate probit:

> We wish to model the bivariate outcomes of whether children attend private 
school and whether the head of the household voted for an increase in property
tax based on the other covariates.

```{r probit-example-3}
school = haven::read_dta('http://www.stata-press.com/data/r13/school.dta')

head(school)

X  = model.matrix(private ~ years + logptax + loginc, school)
y1 = school$private
y2 = school$vote

init = c(0, rep(0, ncol(X)*2))

# you'll probably get a warning or two, ignore; takes a couple seconds
fit = optim(
  fn  = bivariate_probit_ll,
  par = init,
  X   = X,
  y1  = y1,
  y2  = y2,
  method = 'BFGS'
)


loglik = fit$value
rho    = fit$par[1]
coefs_private = fit$par[2:(ncol(X) + 1)]
coefs_vote = fit$par[(ncol(X) + 2):length(init)]
names(coefs_private) = names(coefs_vote) = c('Int', 'years', 'logptax', 'loginc')

list(
  loglik  = loglik,
  rho     = rho,
  Private = coefs_private,
  Vote    = coefs_vote
)
```




## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/bivariateProbit.R
