
# Quantile Regression

## Data Setup

We'll use the <span class="pack" style = "">quantreg</span> package for comparison, and the classic data set on Belgian household income and food expenditure.  Scale income if you want a meaningful 'centercept'.

```{r qr-setup}
library(tidyverse)

library(quantreg)

data(engel)          
# engel$income = scale(engel$income)

X = cbind(1, engel$income)
colnames(X) = c('Intercept', 'income')
```

## Function

Loss function.  It really is this simple.

```{r qr-func}
qreg <- function(par, X, y, tau) {
  lp = X%*%par
  res = y - lp
  loss = ifelse(res < 0 , -(1 - tau)*res, tau*res)
  sum(loss)
}
```

## Estimation

We'll estimate the median to start. Compare `optim` output with <span class="pack" style = "">quantreg</span> package.

```{r qr-est}
optim(
  par = c(intercept = 0, income = 0),
  fn  = qreg,
  X   = X,
  y   = engel$foodexp,
  tau = .5
)$par

rq(foodexp ~ income, tau = .5, data = engel)
```


### Other quantiles

Now we will add additional quantiles to estimate.

```{r qr-quants}
# quantiles
qs = c(.05, .1, .25, .5, .75, .9, .95)

fit_rq = coef(rq(foodexp ~ income, tau = qs, data = engel))

fit_qreg = map_df(qs, function(tau)
  data.frame(t(
    optim(
      par = c(intercept = 0, income = 0),
      fn  = qreg,
      X   = X,
      y   = engel$foodexp,
      tau = tau
    )$par
  )))
```


## Comparison

Compare results.

```{r qr-compare, echo =FALSE}
rbind(fit_rq, t(fit_qreg)) %>% 
  as.data.frame() %>% 
  rownames_to_column('coef') %>% 
  mutate(` ` = rep(c('fit_rq', 'fit_qreg'), each = 2)) %>% 
  select(` `, everything()) %>% 
  kable_df()
```


## Visualize

Let's visualize the results.

```{r qr-vis, echo=FALSE}
engel %>%
  qplot(data = .,
        income,
        foodexp,
        color = I(scales::alpha('orange', .25))) +
  geom_abline(aes(
    intercept = intercept,
    slope = income,
    color = group
  ),
  data = data.frame(fit_qreg, group = factor(qs))) +
  scico::scale_color_scico_d(palette = 'vik')
```


## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/quantile_regression.Rmd
