
# Quantile Regression



## Loss function

```{r qreg}
qreg = function(par, X, y, tau) {
  lp = X%*%par
  res = y - lp
  loss = ifelse(res < 0 , -(1-tau)*res, tau*res)
  sum(loss)
}
```



### Data Prep

We'll use the <span class="pack" style = "">quantreg</span> package for comparison, and the classic data set on Belgian household income and food expenditure.  Scale income if you want a meaningful 'centercept'.

```{r qr-setup}
library(quantreg)
data(engel)          
### engel$income = scale(engel$income)
```

### Median estimation

Compare `optim` output with <span class="pack" style = "">quantreg</span> package.

```{r qr-est}
optim(
  par = c(0, 0),
  fn  = qreg,
  X   = cbind(1, engel$income),
  y   = engel$foodexp,
  tau = .5
)$par

rq(foodexp ~ income, tau = .5, data = engel)
```


### Other quantiles

Now we will add additional quantiles to estimate.

```{r qreg-quants}
# quantiles
qs = c(.05, .1, .25, .5, .75, .9, .95)

resrq = coef(rq(foodexp ~ income, tau = qs, data = engel))


resoptim = map_df(qs, function(tau)
  data.frame(t(
    optim(
      par = c(0, 0),
      fn  = qreg,
      X   = cbind(1, engel$income),
      y   = engel$foodexp,
      tau = tau
    )$par
  )))


# compare results

rbind(resrq, t(resoptim)) %>% round(2)
```

### Visualize

Let's visualize the results.

```{r, results='hide'}
engel %>%
  qplot(data = .,
        income,
        foodexp,
        color = I(scales::alpha('orange', .25))) +
  geom_abline(aes(
    intercept = X1,
    slope = X2,
    color = group
  ),
  data = data.frame(resoptim, group = factor(qs)))
```


## Source

Base R source code found at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/quantile_regression.Rmd