# L2 (ridge) regularization


Compare to the [lasso chapter][L1 (lasso) regularization].  A more conceptual depiction of the both can be found in the [penalized ML chapter][Penalized Maximum Likelihood].

## Data Setup

```{r ridge-setup}
library(tidyverse)

set.seed(8675309)

N = 500
p = 10
X = scale(matrix(rnorm(N * p), ncol = p))
b = c(.5, -.5, .25, -.25, .125, -.125, rep(0, 4))
y = scale(X %*% b + rnorm(N, sd = .5))
```



## Function

```{r ridge-function}
ridge <- function(w, X, y, lambda = .1) {
  # X: model matrix; 
  # y: target; 
  # lambda: penalty parameter; 
  # w: the weights/coefficients
  
  crossprod(y - X %*% w) + lambda * length(y) * crossprod(w)
}
```



## Estimation

Note, if `lambda = 0`, i.e. no penalty, the result is the same as what you would get from the base R  <span class="func" style = "">lm.fit</span>.

```{r ridge-est}
fit_ridge = optim(
  rep(0, ncol(X)),
  ridge,
  X = X,
  y = y,
  lambda = .1,
  method = 'BFGS'
)
```


Analytical result.

```{r ridge-analytical}
fit_ridge2 = solve(crossprod(X) + diag(length(y)*.1, ncol(X))) %*% crossprod(X, y)
```


An alternative approach using 'augmented' data (note `sigma` is ignored as it equals 1, but otherwise
X/sigma and y/sigma).


```{r ridge-augmented}
X2 = rbind(X, diag(sqrt(length(y)*.1), ncol(X)))
y2 = c(y, rep(0, ncol(X)))

tail(X2)
tail(y2)

fit_ridge3 = solve(crossprod(X2)) %*% crossprod(X2, y2)
```

The <span class="pack" style = "">glmnet</span> approach is by default a mixture
of ridge and lasso penalties, setting `alpha = 1` reduces to lasso (`alpha = 0`
would be ridge). We set the lambda to a couple values while only wanting the one
set to the same lambda value as above (`s`).

```{r ridge-glmnet}
library(glmnet)

fit_glmnet = coef(
  glmnet(
    X,
    y,
    alpha = 0,
    lambda = c(10, 1, .1),
    thresh = 1e-12,
    intercept = F
  ), 
  s = .1
)

```


## Comparison

We can now compare the coefficients of all our results.

```{r ridge-compare, echo=FALSE}
data.frame(
  lm     = coef(lm(y ~ . - 1, data.frame(X))),
  ridge  = fit_ridge$par,
  ridge2 = fit_ridge2,
  ridge3 = fit_ridge3,
  glmnet = fit_glmnet[-1, 1],
  truth  = b
) %>% 
  kable_df()
```



## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/ridge.R
