# Stochastic Gradient Descent

Here we have 'online' learning via stochastic gradient descent.  See the
[standard gradient descent][Gradient Descent] chapter. In the following, we have
basic data for standard regression, but in this 'online' learning case, we can
assume each observation comes to us as a stream over time rather than as a
single batch, and would continue coming in.  Note that there are plenty of
variations of this, and it can be applied in the batch case as well.  Currently
no stopping point is implemented in order to trace results over all data
points/iterations. On revisiting this much later, I thought it useful to add
that I believe this was motivated by the example in Murphy's Probabilistic
Machine Learning text.

## Data Setup

Create some data for a standard linear regression.

```{r sgd-setup}
library(tidyverse)

set.seed(1234)

n  = 1000
x1 = rnorm(n)
x2 = rnorm(n)
y  = 1 + .5*x1 + .2*x2 + rnorm(n)
X  = cbind(Intercept = 1, x1, x2)
```




## Function

The estimating function using the *adagrad* approach.

```{r sgd}
sgd <- function(
  par,                       # parameter estimates
  X,                         # model matrix
  y,                         # target variable
  stepsize = 1,              # the learning rate
  stepsize_tau = 0,          # if > 0, a check on the LR at early iterations
  average = FALSE            # a variation of the approach
){
  
  # initialize
  beta = par
  names(beta) = colnames(X)
  betamat = matrix(0, nrow(X), ncol = length(beta))      # Collect all estimates
  fits = NA                                      # Collect fitted values at each point
  loss = NA                                      # Collect loss at each point
  s = 0                                          # adagrad per parameter learning rate adjustment
  eps  = 1e-8                                    # a smoothing term to avoid division by zero
  
  for (i in 1:nrow(X)) {
    Xi   = X[i, , drop = FALSE]
    yi   = y[i]
    LP   = Xi %*% beta                           # matrix operations not necessary, 
    grad = t(Xi) %*% (LP - yi)                   # but makes consistent with standard gd func
    s    = s + grad^2                            # adagrad approach
    
    # update
    beta = beta - stepsize/(stepsize_tau + sqrt(s + eps)) * grad 

    if (average & i > 1) {
      beta =  beta - 1/i * (betamat[i - 1, ] - beta)          # a variation
    } 
    
    betamat[i,] = beta
    fits[i]     = LP
    loss[i]     = (LP - yi)^2
    grad_old = grad
  }
  
  LP = X %*% beta
  lastloss = crossprod(LP - y)
  
  list(
    par    = beta,                               # final estimates
    par_chain = betamat,                         # estimates at each iteration
    RMSE   = sqrt(sum(lastloss)/nrow(X)),
    fitted = LP
  )
}
```



## Estimation


Set starting values.

```{r sgd-start}
starting_values = rep(0, 3)
```

For any particular data you might have to fiddle with the `stepsize`, perhaps
choosing one based on cross-validation with old data.

```{r sgd-est}
fit_sgd = sgd(
  starting_values,
  X = X,
  y = y,
  stepsize     = .1,
  stepsize_tau = .5,
  average = FALSE
)

str(fit_sgd)

fit_sgd$par
```


## Comparison

We can compare to standard linear regression.

```{r sgd-compare-lm}
# summary(lm(y ~ x1 + x2))
coef1 = coef(lm(y ~ x1 + x2))
```

```{r sgd-compare-lm-show, echo=FALSE}
rbind(
  fit_sgd = fit_sgd$par[, 1],
  lm = coef1
) %>% 
  kable_df()
```




## Visualize Estimates


```{r sgd-visualize, echo=FALSE}
library(tidyverse)

gd = data.frame(fit_sgd$par_chain) %>% 
  mutate(Iteration = 1:n())

gd = gd %>%
  pivot_longer(cols = -Iteration,
               names_to = 'Parameter',
               values_to = 'Value') %>%
  mutate(Parameter = factor(Parameter, labels = colnames(X)))

ggplot(aes(
  x = Iteration,
  y = Value,
  group = Parameter,
  color = Parameter
),
data = gd) +
  geom_path() +
  geom_point(data = filter(gd, Iteration == n), size = 3) +
  geom_text(
    aes(label = round(Value, 2)),
    hjust = -.5,
    angle = 45,
    size  = 4,
    data  = filter(gd, Iteration == n)
  ) + 
  scico::scale_color_scico_d(end = .75)
```





## Data Set Shift

This data includes a shift of the previous data, where the data fundamentally changes at certain times.

### Data Setup

We'll add data with different underlying generating processes.

```{r sgd-shift-setup}
set.seed(1234)

n2   = 1000
x1.2 = rnorm(n2)
x2.2 = rnorm(n2)
y2 = -1 + .25*x1.2 - .25*x2.2 + rnorm(n2)
X2 = rbind(X, cbind(1, x1.2, x2.2))
coef2 = coef(lm(y2 ~ x1.2 + x2.2))
y2 = c(y, y2)

n3    = 1000
x1.3  = rnorm(n3)
x2.3  = rnorm(n3)
y3    = 1 - .25*x1.3 + .25*x2.3 + rnorm(n3)
coef3 = coef(lm(y3 ~ x1.3 + x2.3))

X3 = rbind(X2, cbind(1, x1.3, x2.3))
y3 = c(y2, y3)
```





### Estimation

We'll use the same function as before.

```{r sgd-est-2}
fit_sgd_shift = sgd(
  starting_values,
  X = X3,
  y = y3,
  stepsize     = 1,
  stepsize_tau = 0,
  average = FALSE
)

str(fit_sgd_shift)
```

### Comparison

Compare with <span class="func" style = "">lm</span> result for each data part.

```{r sgd-compare-2, echo=F}
lm_coef = rbind(lm_part1 = coef1, lm_part2 = coef2, lm_part3 = coef3) %>% 
  data.frame() %>% 
  rename(Intercept = X.Intercept.)

sgd_coef = fit_sgd_shift$par_chain[c(n, n + n2, n + n2 + n3), ] %>% 
  data.frame()

rownames(sgd_coef) = c('sgd_part1','sgd_part2','sgd_part3')
colnames(sgd_coef) = colnames(lm_coef)

bind_rows(lm_coef, sgd_coef) %>% 
  kable_df()
```


### Visualize Estimates

Visualize estimates across iterations.

```{r sgd-visualize-2, echo=FALSE}
gd = data.frame(fit_sgd_shift$par_chain) %>% 
  mutate(Iteration = 1:n())

gd = gd %>% 
  pivot_longer(cols = -Iteration,
               names_to = 'Parameter', 
               values_to = 'Value') %>% 
  mutate(Parameter = factor(Parameter, labels = colnames(X)))


ggplot(aes(x = Iteration,
           y = Value,
           group = Parameter,
           color = Parameter
           ),
       data = gd) +
  geom_path() +
  geom_point(data = filter(gd, Iteration %in% c(n, n + n2, n + n2 + n3)),
             size = 3) +
  geom_text(
    aes(label = round(Value, 2)),
    hjust = -.5,
    angle = 45,
    data = filter(gd, Iteration %in% c(n, n + n2, n + n2 + n3)),
    size = 4,
    show.legend = FALSE
  ) + 
  scico::scale_color_scico_d(end = .75, alpha = .5)
```


## SGD Variants

The above uses the *Adagrad* approach for stochastic gradient descent, but there are many variations.  A good resource can be found [here](https://ruder.io/optimizing-gradient-descent/), as well as this [post covering more recent developments](https://johnchenresearch.github.io/demon/).  We will compare the *Adagrad*, *RMSprop*, *Adam*, and *Nadam* approaches.



### Data Setup

For this demo we'll bump the sample size. I've also made the coefficients a little different.

```{r sgd-variant-setup, cache.rebuild=F}
library(tidyverse)

set.seed(1234)

n  = 10000
x1 = rnorm(n)
x2 = rnorm(n)
X  = cbind(Intercept = 1, x1, x2)
true = c(Intercept = 1, x1 = 1, x2 = -.75)

y  = X %*% true + rnorm(n)
```


### Function

For this we'll add a functional component to the primary function.  We create a [function factory](https://adv-r.hadley.nz/function-factories.html) `update_ff` that, based on the input will create an appropriate update step (`update`) for use each iteration.  This is mostly is just a programming exercise, but might allow you to add additional components arguments or methods more easily.

```{r sgd-variant-func}
sgd <- function(
  par,                       # parameter estimates
  X,                         # model matrix
  y,                         # target variable
  stepsize = 1e-2,           # the learning rate; suggest 1e-3 for non-adagrad methods
  type = 'adagrad',          # one of adagrad, rmsprop, adam or nadam
  average = FALSE,           # a variation of the approach
  ...                        # arguments to pass to an updating function, e.g. gamma in rmsprop
){
  
  # initialize
  beta = par
  names(beta) = colnames(X)
  betamat = matrix(0, nrow(X), ncol = length(beta))      # Collect all estimates
  v    = rep(0, length(beta))                    # gradient variance (sum of squares)
  m    = rep(0, length(beta))                    # average of gradients for n/adam
  eps  = 1e-8                                    # a smoothing term to avoid division by zero
  grad_old = rep(0, length(beta))
  
  update_ff <- function(type, ...) {
    
    # if stepsize_tau > 0, a check on the LR at early iterations
    adagrad <- function(grad, stepsize_tau = 0) {
      v <<- v + grad^2  
      
      stepsize/(stepsize_tau + sqrt(v + eps)) * grad
    }
    
    rmsprop <- function(grad, grad_old, gamma = .9) {
      v = gamma * grad_old^2 + (1 - gamma) * grad^2
      
      stepsize / sqrt(v + eps) * grad
    }
    
    adam <- function(grad, b1 = .9, b2 = .999) {
      m <<- b1 * m + (1 - b1) * grad
      v <<- b2 * v + (1 - b2) * grad^2
      
      if (type == 'adam')
        # dividing v and m by 1 - b*^i is the 'bias correction'
        stepsize/(sqrt(v / (1 - b2^i)) + eps) *  (m / (1 - b1^i))
      else 
        # nadam
        stepsize/(sqrt(v / (1 - b2^i)) + eps) *  (b1 * m  +  (1 - b1)/(1 - b1^i) * grad)
    }
    
    switch(
      type,
      adagrad = function(grad, ...) adagrad(grad, ...),
      rmsprop = function(grad, ...) rmsprop(grad, grad_old, ...),
      adam    = function(grad, ...) adam(grad, ...),
      nadam   = function(grad, ...) adam(grad, ...)
    )
  }

  update = update_ff(type, ...)
  
  for (i in 1:nrow(X)) {
    Xi   = X[i, , drop = FALSE]
    yi   = y[i]
    LP   = Xi %*% beta                           # matrix operations not necessary, 
    grad = t(Xi) %*% (LP - yi)                   # but makes consistent with standard gd func
    
    # update
    beta = beta - update(grad, ...)
    
    if (average & i > 1) {
      beta = beta - 1/i * (betamat[i - 1, ] - beta)   # a variation
    } 
    
    betamat[i,] = beta
    grad_old = grad
  }
  
  LP = X %*% beta
  lastloss = crossprod(LP - y)
  
  list(
    par = beta,                               # final estimates
    par_chain = betamat,                      # estimates at each iteration
    RMSE = sqrt(sum(lastloss)/nrow(X)),
    fitted = LP
  )
}
```

### Estimation

We'll now use all four methods for estimation.

```{r sgd-variant-est, cache.rebuild=F}
starting_values = rep(0, ncol(X))
# starting_values = runif(3, min = -1)

fit_adagrad = sgd(
  starting_values,
  X = X,
  y = y,
  stepsize = .1  # suggestion is .01 for many settings, but this works better here
)

fit_rmsprop = sgd(
  starting_values,
  X = X,
  y = y,
  stepsize = 1e-3,
  type = 'rmsprop'
)

fit_adam = sgd(
  starting_values,
  X = X,
  y = y,
  stepsize = 1e-3,
  type = 'adam'
)

fit_nadam = sgd(
  starting_values,
  X = X,
  y = y,
  stepsize = 1e-3,
  type = 'nadam'
)
```

### Comparison

We'll compare our results to standard linear regression and the true values.

```{r sgd-variant-compare, echo=FALSE, cache.rebuild=F}
init = map_df(
  list(
    fit_adagrad = fit_adagrad,
    fit_rmsprop = fit_rmsprop,
    fit_adam    = fit_adam,
    fit_nadam   = fit_nadam
  ), 
  function(x) data.frame(t(x$par)), 
  .id = 'fit'
)

fit_lm = data.frame(fit = 'fit_lm', t(lm.fit(X, y)$coef))
true = data.frame(fit = 'true', t(c(Intercept = 1, x1 = 1, x2 = -.75)))

rbind(init, fit_lm, true) %>% 
  kable_df(digits = 4)
```


### Visualize Estimates

We can visualize the route of estimation for each technique. While Adagrad works well for this particular problem, in standard machine learning contexts with possibly millions of parameters, and possibly massive data, it would quickly get to a point where it is no longer updating (the denominator continues to grow).  These other techniques are attempts to get around the limitations of Adagrad. 


```{r sgd-variant-vis, echo=F}
bind_rows(
  fit_adagrad = data.frame(fit_adagrad$par_chain),
  fit_rmsprop = data.frame(fit_rmsprop$par_chain),
  fit_adam = data.frame(fit_adam$par_chain),
  fit_nadam = data.frame(fit_nadam$par_chain),
  .id = 'fit'
) %>% 
  rename(Intercept = X1, x1 = X2, x2 = X3) %>% 
  mutate(iter = rep(1:n, 4)) %>% 
  pivot_longer(-c(iter, fit), names_to = 'parameter') %>% 
  ggplot(aes(iter, value)) +
  geom_line(aes(color = parameter)) +
  facet_wrap(~fct_inorder(fit)) +
  scico::scale_color_scico_d(end = .8)
```


```{r sgd-variant-vis-shift, eval=FALSE, echo=F}
fit_sgd_shift_ada = sgd(
  starting_values,
  X = X3,
  y = y3,
  stepsize = .1, 
  average = T
)

fit_sgd_shift_nadam = sgd(
  starting_values,
  X = X3,
  y = y3,
  stepsize = 1e-3,
  type = 'nadam', 
  average = T
)

bind_rows(
  fit_shift_adagrad = data.frame(fit_sgd_shift_ada$par_chain),
  fit_shift_nadam = data.frame(fit_sgd_shift_nadam$par_chain),
  .id = 'fit'
) %>% 
  rename(Intercept = X1, x1 = X2, x2 = X3) %>% 
  mutate(iter = rep(1:nrow(X3), 2)) %>% 
  pivot_longer(-c(iter, fit), names_to = 'parameter') %>% 
  ggplot(aes(iter, value)) +
  geom_line(aes(color = parameter)) +
  facet_wrap(~fct_inorder(fit)) +
  scico::scale_color_scico_d(end = .8)
```




## Source

Original code available at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/stochastic_gradient_descent.R
