# (PART\*)  Other {-}

```{r suppsetup, include = FALSE, cache=FALSE}
# pretty much nothing can be done to get bookdown to use the correct environment
# library(reticulate)
# use_condaenv('r-reticulate')
knitr::opts_chunk$set(eval = F)
```


# Supplemental

## Other Languages

When doing some of these models and algorithms, I had some other code to work with in another language, or, at the time, just wanted to try it in that language. There is not a whole lot here, but it still may be useful to some.  Refer to the corresponding chapter of R code for context.

### Python Demos

#### Linear Regression {#python-linreg}

##### Data Setup




```{python py-linreg-setup, eval =F}
import numpy as np
import pandas as pd
from scipy.stats import norm
from scipy.optimize import minimize

np.random.seed(123) # ensures replication

# predictors and response
# increasing N will get estimated values closer to the known parameters

N = 1000 # sample size
k = 2   # number of desired predictors
X = np.matrix(np.random.normal(size = N * k)).reshape(N, k)
y = -.5 + .2*X[:, 0] + .1*X[:, 1] + np.random.normal(scale = .5, size = N).reshape(N, 1)

dfXy = pd.DataFrame(np.column_stack([X, y]), columns = ['X1', 'X2', 'y'])
```

##### Functions

A maximum likelihood approach.

```{python py-linreg-ml, eval=FALSE}
def lm_ml(par, X, y):
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  
  # setup
  beta   = par[1:].reshape(X.shape[1], 1)    # coefficients
  sigma  = par[0]                            # error sd
  # N = X.shape[0]
  
  # linear predictor
  LP = X * beta                              # linear predictor
  mu = LP                                    # identity link in the glm sense
  
  # calculate likelihood

  L = norm.logpdf(y, loc = mu, scale = sigma) # log likelihood; or use norm.logpdf
  
  # L =  -.5*N*log(sigma2) - .5*(1/sigma2)*crossprod(y-mu)    # alternate log likelihood form

  L = -np.sum(L)       # optim by default is minimization, and we want to maximize the likelihood 
  
  return(L)
```

An approach via least squares loss function.

```{python py-linreg-ls, eval=FALSE}
def lm_ls(par, X, y):
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  
  # setup
  beta = par.reshape(3, 1)              # coefficients
  N = X.shape[0]
  p = X.shape[1]
  
  # linear predictor
  LP = X * beta                              # linear predictor
  mu = LP                                    # identity link in the glm sense
  # 
  # # squared error loss
  # 
  return(np.sum(np.square(y - mu)))
```


##### Estimation

```{python linreg-est, eval=FALSE}
X_mm = np.column_stack([np.repeat(1, N).reshape(N, 1), X])

# you may get warnings, they can be ignored
fit_ml = minimize(
  fun = lm_ml,
  x0  = [1, 0, 0, 0],
  args    = (X_mm, y),
  bounds  = ((0, None), (None, None), (None, None), (None, None)),
  method  = 'L-BFGS-B', 
  tol     = 1e-12, 
  options = {'maxiter': 500}
)


# can use least_squares directly
# from scipy.optimize import least_squares

fit_ls = minimize(
  lm_ls,
  x0   = np.array([0, 0, 0]),
  args = (X_mm, y), 
  tol     = 1e-12, 
  options = {'maxiter': 500}
)
```

##### Comparison

```{python py-linreg-compare, eval=FALSE}
import statsmodels.formula.api as smf

model_sm_ols = smf.ols('y ~ X1 + X2', data = dfXy)
fit_sm_ols   = model_sm_ols.fit()

pd.DataFrame(
  [
  np.append(fit_ml.x[0]**2, fit_ml.x[1:]), 
  np.append(fit_ls.fun/(N - X_mm.shape[1] - 1), fit_ls.x), 
  np.append(fit_sm_ols.scale, fit_sm_ols.params)
  ], 
  columns = ['sigma','Int', 'b_X1', 'b_X2'], 
  index   = ['ML', 'OLS', 'SM']
  )
```

```{r py-linreg-results, echo=FALSE}
cat("
        sigma       Int      b_X1    b_X2
ML   0.240332 -0.494647  0.213036  0.0815
OLS  0.241297 -0.494647  0.213036  0.0815
SM   0.241055 -0.494647  0.213036  0.0815    
    ")
```



#### Logistic Regression {#python-logreg}

##### Data Setup

```{python py-logreg-setup, eval=F}
import numpy as np
import pandas as pd
from scipy.stats import logistic, binom
from scipy.optimize import minimize

np.random.seed(123) # ensures replication

# predictors and response
# increasing N will get estimated values closer to the known parameters

N = 2500 # sample size
k = 2   # number of desired predictors
X = np.matrix(np.random.normal(size = N * k)).reshape(N, k) 
eta = -.5 + .2*X[:, 0] + .1*X[:, 1]

pi = 1/(1 + np.exp(-eta))

y = np.random.binomial(1, p = pi, size = (N, 1))

dfXy = pd.DataFrame(np.column_stack([X, y]), columns = ['X1', 'X2', 'y'])
```

##### Functions 
 
A maximum likelihood approach.

```{python py-logreg_ml, eval=F}
def logreg_ml(par, X, y):
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: response
  
  # setup
  beta   = par.reshape(X.shape[1], 1)             # coefficients
  # N = X.shape[0]
  
  # linear predictor
  LP = np.dot(X, beta)                              # linear predictor
  mu = 1/(1 + np.exp(-LP))                   # logit link
  
  # calculate likelihood
  L = binom.logpmf(y, 1, mu) # log likelihood
  
  # L =  np.multiply(y, np.log(mu)) + np.multiply(1 - y, np.log(1 - mu))    # alternate log likelihood form

  L = -np.sum(L)       # optim by default is minimization, and we want to maximize the likelihood 
  
  return(L)
```


Another approach via exponential loss function.

```{python py-logreg_exp, eval=F}
def logreg_exp(par, X, y):
  # Arguments
  # par: parameters to be estimated
  # X: predictor matrix with intercept column
  # y: target
  
  # setup
  beta = par.reshape(X.shape[1], 1)                                    # coefficients
  
  # linear predictor
  LP = np.dot(X, beta)                              # linear predictor

  # calculate exponential loss function (convert y to -1:1 from 0:1)
  L = np.sum(np.exp(-np.multiply(np.where(y, 1, -1), LP) * .5))
  
  return(L)
```




##### Estimation

Setup for use with <span class="func" style = "">optim</span>.

```{python py-logreg-est, eval=F}
X_mm = np.column_stack([np.repeat(1, N).reshape(N, 1), X])

fit_ml = minimize(
  fun = logreg_ml,
  x0  = [0, 0, 0],
  args    = (X_mm, y),
  method  = 'BFGS', 
  tol     = 1e-12, 
  options = {'maxiter': 500}
)

fit_exp = minimize(
  fun = logreg_exp,
  x0  = [0, 0, 0],
  args    = (X_mm, y),
  method  = 'BFGS', 
  tol     = 1e-12, 
  options = {'maxiter': 500}
)

pars_ml  = fit_ml.x
pars_exp = fit_exp.x
```

##### Comparison

```{python py-logreg-compare, eval=FALSE}
import statsmodels.formula.api as smf

model_sm_glm = smf.logit('y ~ X1 + X2', data = dfXy)
fit_sm_glm   = model_sm_glm.fit()

pd.DataFrame(
  [
  pars_ml, 
  pars_exp, 
  fit_sm_glm.params
  ], 
  columns = ['Int', 'b_X1', 'b_X2'], 
  index   = ['fit_ml', 'fit_exp', 'SM']
  )
```

```{r py-logreg-results, echo=FALSE}
cat("
              Int      b_X1      b_X2
fit_ml  -0.505637  0.234567  0.054700
fit_exp -0.504802  0.231859  0.053134
SM      -0.505637  0.234567  0.054700  
    ")
```



#### Quantile Regression {#python-qreg}

See the [R chapter][Quantile Regression] for details.

##### Data Setup

```{python py-qreg-setup, eval=FALSE}
import pandas as pd
import numpy as np

# originally from quantreg package
engel = pd.read_csv('data/engel.csv')

X = np.column_stack([
  np.ones((engel.shape[0],1)),
  np.asarray(engel['income'])
  ])
  
y = np.asarray(engel['foodexp'])
```



```{python py-qreg-func, eval = F}
def qreg(par, X, y, tau):
  lp = np.dot(X, par)
  res = y - lp
  loss = np.where(res < 0 , -(1 - tau)*res, tau*res)
  
  return(np.sum(loss))
```



##### Estimation

We'll estimate the median to start. 


```{python py-qreg-est, eval=FALSE}
# from scipy.stats import logistic, binom
from scipy.optimize import minimize


minimize(
  fun = qreg,
  x0  = [0, 0],
  args    = (X, y, .5),
  method  = 'BFGS', 
  tol     = 1e-12, 
  options = {'maxiter': 500}
)
```

##### Other quantiles

Now we will add additional quantiles to estimate.

```{python py-qreg-quants}
pandas as pd

# quantiles
qs = [.05, .1, .25, .5, .75, .9, .95]
fit = []

for tau in qs:
  init = minimize(
    fun = qreg,
    x0  = [0, 0],
    args    = (X, y, tau),
    method  = 'BFGS', 
    tol     = 1e-12, 
    options = {'maxiter': 500}
  )
  
  fit.append(
    pd.DataFrame(init.x.reshape(1,2),  columns = ['int', 'slope'])
  )
    
fit_qr = pd.concat(fit)


from plotnine import *

(
  ggplot(aes(x = 'income', y = 'foodexp'), data = engel) +
    geom_point() +
    geom_abline(aes(intercept = 'int', slope = 'slope', color = 'int'), 
    data = fit_qr)
)
```




#### Nelder-Mead {#python-nelder}

```{python py-nelder-mead, eval = FALSE}
import copy

'''
    Francois Chollet's Nelder-Mead in Python.
    https://github.com/fchollet/nelder-mead/blob/master/nelder_mead.py
    Pure Python/Numpy implementation of the Nelder-Mead algorithm.
    Reference: https://en.wikipedia.org/wiki/Nelder%E2%80%93Mead_method
'''


def nelder_mead(
  f,
  x_start,
  step = 0.1,
  no_improve_thr  = 10e-6,
  no_improv_break = 10,
  max_iter = 0,
  alpha = 1.,
  gamma = 2.,
  rho   = 0.5,
  sigma = 0.5
):
    '''
        @param f (function): function to optimize, must return a scalar score
            and operate over a numpy array of the same dimensions as x_start
        @param x_start (numpy array): initial position
        @param step (float): look-around radius in initial step
        @no_improv_thr,  no_improv_break (float, int): break after no_improv_break iterations with
            an improvement lower than no_improv_thr
        @max_iter (int): always break after this number of iterations.
            Set it to 0 to loop indefinitely.
        @alpha, gamma, rho, sigma (floats): parameters of the algorithm
            (see Wikipedia page for reference)
        return: tuple (best parameter array, best score)
    '''

    # init
    dim = len(x_start)
    prev_best = f(x_start)
    no_improv = 0
    res = [[x_start, prev_best]]

    for i in range(dim):
        x = copy.copy(x_start)
        x[i] = x[i] + step
        score = f(x)
        res.append([x, score])

    # simplex iter
    iters = 0
    while 1:
        # order
        res.sort(key=lambda x: x[1])
        best = res[0][1]

        # break after max_iter
        if max_iter and iters >= max_iter:
            return res[0]
        iters += 1

        # break after no_improv_break iterations with no improvement
        if iters//10 == 0:
          print('...best so far:', best)

        if best < prev_best - no_improve_thr:
            no_improv = 0
            prev_best = best
        else:
            no_improv += 1

        if no_improv >= no_improv_break:
            return res[0]

        # centroid
        x0 = [0.] * dim
        for tup in res[:-1]:
            for i, c in enumerate(tup[0]):
                x0[i] += c / (len(res)-1)

        # reflection
        xr = x0 + alpha*(x0 - res[-1][0])
        rscore = f(xr)
        if res[0][1] <= rscore < res[-2][1]:
            del res[-1]
            res.append([xr, rscore])
            continue

        # expansion
        if rscore < res[0][1]:
            xe = x0 + gamma*(x0 - res[-1][0])
            escore = f(xe)
            if escore < rscore:
                del res[-1]
                res.append([xe, escore])
                continue
            else:
                del res[-1]
                res.append([xr, rscore])
                continue

        # contraction
        xc = x0 + rho*(x0 - res[-1][0])
        cscore = f(xc)
        if cscore < res[-1][1]:
            del res[-1]
            res.append([xc, cscore])
            continue

        # reduction
        x1 = res[0][0]
        nres = []
        for tup in res:
            redx = x1 + sigma*(tup[0] - x1)
            score = f(redx)
            nres.append([redx, score])
        res = nres


if __name__ == "__main__":
    # test
    import math
    import numpy as np

    def f(x):
        return math.sin(x[0]) * math.cos(x[1]) * (1. / (abs(x[2]) + 1))

nelder_mead(f, np.array([0., 0., 0.]))
```



#### HMM {#python-hmm}

```{python py-hmm, eval = FALSE}
#!/usr/bin/env python3
# -*- coding: utf-8 -*-
"""
From the wikipedia page with slight modification
https://en.wikipedia.org/wiki/Viterbi_algorithm#Example
"""

def viterbi(obs, states, start_p, trans_p, emit_p):
    V = [{}]

    for st in states:
        V[0][st] = {"prob": start_p[st] * emit_p[st][obs[0]], "prev": None}

    # Run Viterbi when t > 0

    for t in range(1, len(obs)):
        V.append({})

        for st in states:
            max_tr_prob = max(V[t-1][prev_st]["prob"]*trans_p[prev_st][st] for prev_st in states)

            for prev_st in states:
                if V[t-1][prev_st]["prob"] * trans_p[prev_st][st] == max_tr_prob:
                    max_prob = max_tr_prob * emit_p[st][obs[t]]
                    V[t][st] = {"prob": max_prob, "prev": prev_st}
                    break

    for line in dptable(V):
        print(line)

    opt = []

    # The highest probability
    max_prob = max(value["prob"] for value in V[-1].values())
    previous = None

    # Get most probable state and its backtrack
    for st, data in V[-1].items():
        if data["prob"] == max_prob:
            opt.append(st)
            previous = st
            break

    # Follow the backtrack till the first observation
    for t in range(len(V) - 2, -1, -1):
        opt.insert(0, V[t + 1][previous]["prev"])
        previous = V[t + 1][previous]["prev"]

    print('The steps of states are ' + ' '.join(opt) + ' with highest probability of %s' % max_prob)


def dptable(V):
    # Print a table of steps from dictionary
    yield " ".join(("%12d" % i) for i in range(len(V)))
    
    for state in V[0]:
        yield "%.7s: " % state + " ".join("%.7s" % ("%f" % v[state]["prob"]) for v in V)

# The function viterbi takes the following arguments: obs is the sequence of
# observations, e.g. ['normal', 'cold', 'dizzy']; states is the set of hidden
# states; start_p is the start probability; trans_p are the transition
# probabilities; and emit_p are the emission probabilities. For simplicity of
# code, we assume that the observation sequence obs is non-empty and that
# trans_p[i][j] and emit_p[i][j] is defined for all states i,j.

# In the running example, the forward/Viterbi algorithm is used as follows:


obs = ('normal', 'cold', 'dizzy')
states = ('Healthy', 'Fever')


start_p = {'Healthy': 0.6, 'Fever': 0.4}

trans_p = {
   'Healthy' : {'Healthy': 0.7, 'Fever': 0.3},
   'Fever' : {'Healthy': 0.4, 'Fever': 0.6}
   }
   
emit_p = {
   'Healthy' : {'normal': 0.5, 'cold': 0.4, 'dizzy': 0.1},
   'Fever' : {'normal': 0.1, 'cold': 0.3, 'dizzy': 0.6}
   }
   
viterbi(obs,states,start_p,trans_p,emit_p)
```




### Julia Demos

I haven't played with Julia in a very long time, but briefly hacked the old code to get something that worked.  As Julia has gone though notable changes, it's doubtful these are very good as far as Julia programming standards go, though conceptually they may still provide some utility.  Perhaps at some point I'll reteach myself the basics and come back to these. In any case the code did run on a [Jupyter notebook](https://jupyter.org/). 

####  Mixed Models

##### One-factor


```{julia jl-one-factor-func, eval = FALSE}
#####################
### Main function ###
#####################

using LinearAlgebra
using Statistics

function one_factor_re_loglike(par::Vector)    
    d, ni = size(y)
    mu = par[1]
    sigma2_mu = par[2]
    sigma2 = par[3]

    Sigmai = sigma2*I(ni) + sigma2_mu*ones(ni, ni)
    l = -(ni*d)/2*log(2*pi) - d/2*log(det(Sigmai))
    
    for i in 1:d
      yi = y[i,:]
      l = l - .5(yi .- mu)' * (Sigmai\(yi .- mu))
    end

    l = -l[1]
    
    return l
end
```


```{julia jl-one-factor-setup, eval = FALSE}
###################
### Data set up ###
###################

y = [22.6 20.5 20.8
     22.6 21.2 20.5
     17.3 16.2 16.6
     21.4 23.7 23.2
     20.9 22.2 22.6
     14.5 10.5 12.3
     20.8 19.1 21.3
     17.4 18.6 18.6
     25.1 24.8 24.9
     14.9 16.3 16.6]


################################
### Starting values and test ###
################################
using Statistics

mu0 = mean(y)
sigma2_mu0 = var(mean(y, dims = 2))
sigma20 = mean(var(y, dims = 2))
theta0  = [mu0, sigma2_mu0, sigma20]

### test
one_factor_re_loglike(theta0)
```


```{julia jl-onefactor-est, eval = FALSE}
###########
### Run ###
###########

using Optim
res = optimize(one_factor_re_loglike, theta0, LBFGS())
res
```

```{r julia-one-factor-res, echo=FALSE, eval=TRUE}
cat(
  "
 * Status: success

 * Candidate solution
    Final objective value:     6.226441e+01

 * Found with
    Algorithm:     L-BFGS

 * Convergence measures
    |x - x'|               = 2.93e-08 ≰ 0.0e+00
    |x - x'|/|x'|          = 1.49e-09 ≰ 0.0e+00
    |f(x) - f(x')|         = 1.42e-14 ≰ 0.0e+00
    |f(x) - f(x')|/|f(x')| = 2.28e-16 ≰ 0.0e+00
    |g(x)|                 = 1.17e-09 ≤ 1.0e-08

 * Work counters
    Seconds run:   1  (vs limit Inf)
    Iterations:    7
    f(x) calls:    22
    ∇f(x) calls:   22
  
  "
)
```


```{julia jl-onefactor-res1, eval = FALSE}
Optim.minimizer(res)
```

```{r julia-onefactor-res1, echo=FALSE, eval=TRUE}
cat("
3-element Array{Float64,1}:
19.599999980440952
12.193999992338886
1.1666666662195693
")
```




```{julia jl-onefactor-res2, eval = FALSE}
Optim.minimum(res)
```

```{r julia-onefactor-res2, echo = FALSE, eval=TRUE}
cat('62.30661224610756')
```


##### Two-factor


```{julia  jl-two-factor-func, eval = FALSE}
using LinearAlgebra
using Statistics

function sfran2_loglike(par::Vector)
    n = length(y)
    mu = par[1]
    
    sigma2_alpha = exp(par[2])
    sigma2_gamma = exp(par[3])
    sigma2 = exp(par[4])

    Sigma = sigma2*I(n) + sigma2_alpha*(Xalpha * Xalpha') + sigma2_gamma * (Xgamma * Xgamma')

    l = -n/2*log(2*pi) - sum(log.(diag(cholesky(Sigma).L))) - .5*(y .- mu)' * (Sigma\(y .- mu))

    l = -l[1]
    return l
end
```


```{julia  jl-two-factor-setup, eval = FALSE}
##################
### Data setup ###
##################
y = [1.39,1.29,1.12,1.16,1.52,1.62,1.88,1.87,1.24,1.18,
     .95,.96,.82,.92,1.18,1.20,1.47,1.41,1.57,1.65]


################################
### Starting values and test ###
################################
yhat = mean(reshape(y, 4, 5), 1)

theta0 = [mean(y), log(var(yhat)), log(var(y)/3), log(var(y)/3)]

sfran2_loglike(theta0)
```





```{julia jl-twofactor-res, eval = FALSE}
###########
### Run ###
###########
using Optim
res = optimize(sfran2_loglike, theta0, method = :l_bfgs)
res
```

```{r julia-twofactor-res0, echo = FALSE, eval=TRUE}
cat(
"
 * Status: success

 * Candidate solution
    Final objective value:     -1.199315e+01

 * Found with
    Algorithm:     L-BFGS

 * Convergence measures
    |x - x'|               = 6.60e-09 ≰ 0.0e+00
    |x - x'|/|x'|          = 1.08e-09 ≰ 0.0e+00
    |f(x) - f(x')|         = 5.33e-15 ≰ 0.0e+00
    |f(x) - f(x')|/|f(x')| = 4.44e-16 ≰ 0.0e+00
    |g(x)|                 = 7.02e-10 ≤ 1.0e-08

 * Work counters
    Seconds run:   0  (vs limit Inf)
    Iterations:    9
    f(x) calls:    23
    ∇f(x) calls:   23

")
```

```{julia jl-twofactor-res1, eval=FALSE}
exp.(Optim.minimizer(res))
```


```{r julia-twofactor-res1, echo = FALSE, eval=TRUE}
cat(
"
4-element Array{Float64,1}:
 3.7434213772629223
 0.053720000000540405
 0.031790000003692476
 0.002290000000530042
"
)
```



```{julia jl-twofactor-res2, eval = FALSE}
-2*Optim.minimum(res)
```

```{r julia-twofactor-res2, echo = FALSE, eval=TRUE}
cat(
  "23.98630759443859"
)
```


### Matlab Demos

I don't code in Matlab, nor have any particular desire to, so this is provided here just for reference.

#### Mixed Models

##### One-factor

```{Matlab ml-one-fac, eval = FALSE}
% matlab from Statistical Modeling and Computation (2014 p 311).  See the 
% associated twofactorRE.R file for details.

function one_factor_re_loglike(mu, sigma2_mu, sigma2, y)
	[d ni] = size(y);
	Sigmai = sigma2*eye(ni) + sigma2_mu*ones(ni,ni);
	l = -(ni*d) / 2*log(2*pi) - d / 2*log(det(Sigmai));
	for i=1:d
	  yi = y(i, :)';
	  l = l - .5*(yi - mu)' * (Sigmai\(yi - mu));
	end
end


y = [22.6 20.5 20.8;
     22.6 21.2 20.5;
     17.3 16.2 16.6;
     21.4 23.7 23.2;
     20.9 22.2 22.6;
     14.5 10.5 12.3;
     20.8 19.1 21.3;
     17.4 18.6 18.6;
     25.1 24.8 24.9;
     14.9 16.3 16.6];


f = @(theta) -one_factor_re_loglike(theta(1), theta(2), theta(3), y);
ybar = mean(y, 2);
theta0 = [mean(ybar) var(ybar) mean(var(y, 0, 2))];
thetahat = fminsearch(f, theta0);
```


##### Two-factor

```{Matlab ml-two-fac, eval = FALSE}
% matlab from Statistical Modeling and Computation (2014 p 314). See the 
% associated twofactorRE.R file for details.

function sfran2_loglike(mu, eta_alpha, eta_gamma, eta, y, Xalpha, Xgamma)
  sigma2_alpha = exp(eta_alpha);
  sigma2_gamma = exp(eta_gamma);
  sigma2 = exp(eta);
  
  n = length(y);
  
  Sigma = sigma2*speye(n) + sigma2_alpha * (Xalpha * Xalpha') + sigma2_gamma * (Xgamma*Xgamma');
  
  l = -n/2 * log(2*pi) - sum(log(diag(chol(Sigma)))) - .5*(y - mu)' * (Sigma\(y - mu));
  
end

y = [1.39 1.29 1.12 1.16 1.52 1.62 1.88 1.87 1.24 1.18 .95 .96 .82 .92 1.18 1.20 1.47 1.41 1.57 1.65];

Xalpha = kron(speye(5),  ones(4,1));

Xgamma = kron(speye(10), ones(2,1));

f = @(theta) -sfran_loglike(theta(1), theta(2), theta(3), theta(4), y, Xalpha, Xgamma);
yhat     = mean(reshape(y, 4, 5));
theta0   = [mean(y) log(var(yhat)) log(var(y)/3) log(var(y)/3)];
thetahat = fminsearch(f, theta0)
```


#### Gaussian Processes {#matlab-gp}

Any updates on the following can be found at the [repo](https://github.com/probml/pmtk3).

```{Matlab ml-gp, eval = FALSE}
function S = gaussSample(arg1, arg2, arg3)
% Returns n samples (in the rows) from a multivariate Gaussian distribution
%
% Examples:
% S = gaussSample(mu, Sigma, 10)
% S = gaussSample(model, 100)
% S = gaussSample(struct('mu',[0], 'Sigma', eye(1)), 3)

% This file is from pmtk3.googlecode.com


switch nargin
    case 3,  mu = arg1; Sigma = arg2; n = arg3;
    case 2, model = arg1; mu = model.mu; Sigma = model.Sigma; n = arg2;
    case 1, model = arg1; mu = model.mu; Sigma = model.Sigma; n = 1; 
    otherwise
        error('bad num args')
end

A = chol(Sigma, 'lower');
Z = randn(length(mu), n);
S = bsxfun(@plus, mu(:), A*Z)';


end
```

```{Matlab ml-gp2, eval = FALSE}
%% Visualize the effect of change the hyper-params for a 1d GP regression
% based on demo_gpr by Carl Rasmussen
%
%% Generate data

% This file is from pmtk3.googlecode.com

n = 20;
rand('state',18);
randn('state',20);
covfunc = {'covSum', {'covSEiso','covNoise'}};
loghyper = [log(1.0); log(1.0); log(0.1)];
x = 15*(rand(n,1)-0.5);
y = chol(feval(covfunc{:}, loghyper, x))'*randn(n,1);        % Cholesky decomp.

xstar = linspace(-7.5, 7.5, 201)';

hyps = [log(1), log(1), log(0.1);...
  log(0.3),log(1.08),log(0.00005);...
  log(3),log(1.16),log(0.89)];

%% compute post pred and plot marginals
for i=1:size(hyps,1)
  loghyper = hyps(i,:)';
  [mu, S2] = gpr(loghyper, covfunc, x, y, xstar);
  S2 = S2 - exp(2*loghyper(3)); % remove observation noise
  
  figure;
  f = [mu+2*sqrt(S2);flipdim(mu-2*sqrt(S2),1)];
  fill([xstar; flipdim(xstar,1)], f, [7 7 7]/8, 'EdgeColor', [7 7 7]/8);
  hold on
  plot(xstar,mu,'k-','LineWidth',2);
  plot(x, y, 'k+', 'MarkerSize', 17);
  axis([-8 8 -3 3])
  printPmtkFigure(sprintf('gprDemoChangeHparams%d', i));
end
```

```{Matlab ml-gp-3, eval = FALSE}
%% Reproduce figure 2.2 from GP book
%
%%

% This file is from pmtk3.googlecode.com

setSeed(0);
L = 1;
xs = (-5:0.2:5)';
ns = length(xs);
keps = 1e-8;
muFn = @(x) 0*x(:).^2;
Kfn = @(x,z) 1*exp(-sq_dist(x'/L,z'/L)/2);


% plot sampled functions from the prior
figure; hold on
for i=1:3
  model = struct('mu', muFn(xs), 'Sigma',  Kfn(xs, xs) + 1e-15*eye(size(xs, 1)));
  fs = gaussSample(model, 1);
  plot(xs, fs, 'k-', 'linewidth', 2)
end
printPmtkFigure('gprDemoNoiseFreePrior')


% generate noise-less training data
Xtrain = [-4, -3, -2, -1, 1]';
ftrain = sin(Xtrain);

% compute posterior predictive
K = Kfn(Xtrain, Xtrain); % K
Ks = Kfn(Xtrain, xs); %K_*
Kss = Kfn(xs, xs) + keps*eye(length(xs)); % K_** (keps is essential!)
Ki = inv(K);
postMu = muFn(xs) + Ks'*Ki*(ftrain - muFn(Xtrain));
postCov = Kss - Ks'*Ki*Ks;

figure; hold on
% plot marginal posterior variance as gray band
mu = postMu(:);
S2 = diag(postCov);
f = [mu+2*sqrt(S2);flipdim(mu-2*sqrt(S2),1)];
fill([xs; flipdim(xs,1)], f, [7 7 7]/8, 'EdgeColor', [7 7 7]/8);

% plot samples from posterior predictive
for i=1:3
  model = struct('mu', postMu(:)', 'Sigma', postCov);
  fs = gaussSample(model, 1);
  plot(xs, fs, 'k-', 'linewidth', 2)
  h=plot(Xtrain, ftrain, 'kx', 'markersize', 12, 'linewidth', 3);
end
printPmtkFigure('gprDemoNoiseFreePost')
```




