# Zero-Inflated Model


Log likelihood function to estimate parameters for a Zero-inflated Poisson model. With examples
and comparison to <span class="pack" style = "">pscl</span> package output. Also includes approach based on Hilbe GLM text.


## Poisson

### Data Setup

Get the data.

```{r zip-setup}
library(tidyverse)

fish = haven::read_dta("http://www.stata-press.com/data/r11/fish.dta")
```

### Function

The log likelihood function.

```{r zip-ll}
ZIP = function(y, X, par) {
  # arguments are response y, predictor matrix X, and parameter named starting points of 'logit' and 'pois'
  
  # Extract parameters
  logitpars = par[grep('logit', names(par))]   
  poispars  = par[grep('pois', names(par))]     
  
  # Logit part; in this function Xlogit = Xpois but one could split X argument into Xlogi and Xpois for example
  Xlogit  = X
  LPlogit = Xlogit %*% logitpars
  logi0   = plogis(LPlogit)  # alternative 1/(1+exp(-LPlogit))
    
  # Poisson part
  Xpois  = X
  mupois = exp(Xpois %*% poispars)
  
  # LLs
  logliklogit = log( logi0 + exp(log(1 - logi0) - mupois) )
  loglikpois  = log(1 - logi0) + dpois(y, lambda = mupois, log = TRUE)
  
  # Hilbe formulation
  # logliklogit = log(logi0 + (1 - logi0)*exp(- mupois) )
  # loglikpois = log(1-logi0) -mupois + log(mupois)*y     #not necessary: - log(gamma(y+1))
    
  y0 = y == 0  # 0 values
  yc = y > 0   # Count part

  loglik = sum(logliklogit[y0]) + sum(loglikpois[yc])
  -loglik
}
```







### Estimation


Get starting values or simply do zeros.

```{r zip-starts}
# for zip: need 'logit', 'pois'
initial_model = glm(
  count ~ persons + livebait,
  data = fish,
  x = TRUE,
  y = TRUE,
  "poisson"
)

# starts = c(logit = coef(initial_model), pois = coef(initial_model))  
starts = c(rep(0, 3), rep(0, 3))

names(starts) = c(paste0('pois.', names(coef(initial_model))),
                  paste0('logit.', names(coef(initial_model))))
```


                                                               

Estimate with <span class="func" style = "">optim</span>.

```{r zip-est}
optPois1 = optim(
  par = starts ,
  fn  = ZIP,
  X   = initial_model$x,
  y   = initial_model$y,
  method  = "BFGS",
  control = list(maxit = 5000, reltol = 1e-12),
  hessian = TRUE
)

# optPois1
```




### Comparison

Extract for clean display.

```{r zip-extract}
B  = optPois1$par
se = sqrt(diag(solve((optPois1$hessian))))
Z  = B/se
p  = pnorm(abs(Z), lower = FALSE)*2
```

Results from <span class="pack" style = "">pscl</span>.

```{r zip-pscl}
library(pscl)

zipoismod = zeroinfl(count ~ persons + livebait, data = fish, dist = "poisson") 
```

Compare.

```{r zip-compare}
summary(zipoismod)$coefficients
round(data.frame(B, se, Z, p), 4)
```




## Negative Binomial

### Function

```{r zinb-ll}
ZINB = function(y, X, par) {
  # arguments are response y, predictor matrix X, and parameter named starting points of 'logit', 'negbin', and 'theta'
  
  # Extract parameters
  logitpars  = par[grep('logit', names(par))]
  negbinpars = par[grep('negbin', names(par))]
  theta = exp(par[grep('theta', names(par))])
  
  # Logit part; in this function Xlogit = Xnegbin but one could split X argument into Xlogit and Xnegbin for example
  Xlogit  = X
  LPlogit = Xlogit %*% logitpars
  logi0   = plogis(LPlogit) 
  
  # Negbin part
  Xnegbin = X
  munb = exp(Xnegbin %*% negbinpars)
  
  # LLs
  logliklogit  = 
    log( 
      logi0 + exp(log(1 - logi0) + suppressWarnings(dnbinom(0, 
                                                            size = theta, 
                                                            mu   = munb, 
                                                            log  = TRUE)))
    )
  
  logliknegbin = log(1 - logi0) + suppressWarnings(dnbinom(y, 
                                                           size = theta, 
                                                           mu   = munb, 
                                                           log  = TRUE))
  
  # Hilbe formulation
  # theta part 
  # alpha = 1/theta  
  # m = 1/alpha
  # p = 1/(1 + alpha*munb)
  
  # logliklogit = log( logi0 + (1 - logi0)*(p^m) )
  # logliknegbin = log(1-logi0) + log(gamma(m+y)) - log(gamma(m)) + m*log(p) + y*log(1-p)   # gamma(y+1) not needed
  
  y0 = y == 0   # 0 values
  yc = y > 0    # Count part
  
  loglik = sum(logliklogit[y0]) + sum(logliknegbin[yc])
  -loglik
}
```


### Estimation

Get starting values or simply do zeros.

```{r zinb-starts}
# for zinb: 'logit', 'negbin', 'theta'
initial_model = model.matrix(count ~ persons + livebait, data = fish) # to get X matrix

startlogi  = glm(count == 0 ~ persons + livebait, data = fish, family = "binomial")
startcount = glm(count ~ persons + livebait, data = fish, family = "poisson")

starts = c(
  negbin = coef(startcount),
  logit = coef(startlogi),
  theta = 1
)  
# starts = c(negbin = rep(0, 3),
#            logit = rep(0, 3),
#            theta = log(1))
```




Estimate with <span class="func" style = "">optim</span>.

```{r zinb-est}
optNB1 = optim(
  par = starts ,
  fn  = ZINB,
  X   = initial_model,
  y   = fish$count,
  method  = "BFGS",
  control = list(maxit = 5000, reltol = 1e-12),
  hessian = TRUE
)
# optNB1
```



### Comparison

Extract for clean display.

```{r zinb-extract}
B  = optNB1$par
se = sqrt(diag(solve((optNB1$hessian))))
Z  = B/se
p  = pnorm(abs(Z), lower = FALSE)*2
```

Results from <span class="pack" style = "">pscl</span>.

```{r zinb-pscl}
# pscl results
library(pscl)

zinbmod1 = zeroinfl(count ~ persons + livebait, data = fish, dist = "negbin")
```

Compare.

```{r zinb-compare}
summary(zinbmod1)
round(data.frame(B, se, Z, p), 4)  # note that theta here is actually log(theta)
```


## Supplemental Example



```{r zinb-supplemental}
data("bioChemists", package = "pscl")

initial_model   = model.matrix(art ~ fem + mar + kid5 + phd + ment, data = bioChemists) # to get X matrix
startlogi  = glm(art==0 ~ fem + mar + kid5 + phd + ment, data = bioChemists, family = "binomial")
startcount = glm(art ~ fem + mar + kid5 + phd + ment, data = bioChemists, family = "quasipoisson")

starts = c(
  negbin = coef(startcount),
  logit  = coef(startlogi),
  theta  = summary(startcount)$dispersion
)  

# starts = c(negbin = rep(0, 6),
#            logit = rep(0, 6),
#            theta = 1)


optNB2 = optim(
  par = starts ,
  fn  = ZINB,
  X   = initial_model,
  y   = bioChemists$art,
  method  = "BFGS",
  control = list(maxit = 5000, reltol = 1e-12),
  hessian = TRUE
)
# optNB2


B  = optNB2$par
se = sqrt(diag(solve((optNB2$hessian))))
Z  = B/se
p  = pnorm(abs(Z), lower = FALSE)*2


library(pscl)
zinbmod = zeroinfl(art ~ . | ., data = bioChemists, dist = "negbin")

summary(zinbmod)$coefficients
round(data.frame(B,se, Z, p), 4)
```



## Source

Original code for ZIP found at
https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/poiszeroinfl.R

Original code for ZINB found at https://github.com/m-clark/Miscellaneous-R-Code/blob/master/ModelFitting/NBzeroinfl.R
